---
title: "MutSeqR: Modelling Mutation Frequencies"
author:
- name: Annette E. Dodge
  affiliation: Environmental Health Science and Research Bureau, Health Canada, Ottawa, ON, Canada.
- name: Matthew J. Meier
  affilitation: Environmental Health Science and Research Bureau, Health Canada, Ottawa, ON, Canada.
  email: matthew.meier@hc-sc.gc.ca
package: MutSeqR
output:
  rmarkdown::html_document:
    toc: true
    toc_float: yes
    code_folding: show
    code_download: yes
    css: custom.css
bibliography: Modelling Mutation Frequencies.bib
---
  
```{r setup, include=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(DT)
library(htmltools)
```

# Load MutSeqR and Example Data

```{r load-lib}
library(MutSeqR)
```

```{r load-ex}
library(ExperimentHub)
# load the index
eh <- ExperimentHub()
```

# Generalized Linear Modeling

An important component of analysing mutagencity data is how
mutation frequency (MF) changes based on experimental variables.

The `model_mf()` function can be used to quantitatively evaluate
how MF changes based on experimental variables. The function models
the proportion of reads that carry a mutation (i.e. the MF) across
user-supplied fixed and/or random effects and interaction parameters.
Depending on the supplied effects, `model_mf()` will automatically
choose to fit either a generalized linear model (GLM) using the `glm()`
function from the `r CRANpkg("stats")` library or a generalized linear mixed model
(GLMM) using the `glmer()` function from the `r CRANpkg("lme4")` library.

## Model Formula
- **Fixed effects**: The experimental variable(s)/factor(s) of interest.
- **Random effects**: Variable used to account for variability within a larger
group. Covariate.
- **Interaction**: The product of two or more interacting fixed_effects. An
interaction implies that the effect of one fixed effect changes depending on
the levels of another fixed effect, indicating a non-additive reationship.
- **Response**: the function will model the MF, taking the mutation count
(`sum`) and the `group_depth` as input.

The model formula is built as:

`cbind(sum, group_depth) ~ fixed_effect1 + fixed_effect2 + ... + (1|random_effect)`

If `test_interaction` is set to TRUE, the model will use the product of
the fixed_effects:

`cbind(sum, group_depth) ~ fixed_effect1 * fixed_effect2 * ... + (1|random_effect)`

## Family
The occurence of mutations is assumed to follow a binomial distribution as:

1. There is a finite number of sequenced bases.
2. A mutation at any given base is equally probable*.
3. Mutations occur independently of other mutations.

* We acknowledge that this does not reflect biological reality.

To account for over-dispersion, the function will fit a GLM with a
quasibinomial distribution.  If the dispersion parameter of the model
is less than 1, the model will be refit using a binomial distribution.
The GLMM will always use a binomial distribution.

The `model_mf` function will fit a generalized linear model to analyse
the effect(s) of given fixed_effects(s) on MF and perform specified
pairwise comparisons between levels of your factors. 

Additional arguments can be passed to the model to further customize it to
your needs. Details on the arguments for the generalized linear model can be
found at
[stats::glm](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/glm)
and for the general linear mixed model at
[lme4::glmer](https://www.rdocumentation.org/packages/lme4/versions/1.1-35.3/topics/glmer). 

## Estimates & Comparisons

`model_mf()` provides estimates of the mean for each level of the fixed
effects. Furthermore, pairwise comparisons can be performed based on a
user-supplied table. Mean estimates and comparisons are conducted using the
doBy R library [@doBy]. Estimates of the back-transformed
standard errors are approximated using the delta method. The p-values are
adjusted for multiple comparisons using the Sidak method. 

## Goodness of Fit

The `model_mf` function will output the Pearson residuals appended to the
`mf_data`. The row with the highest residual will be printed to the console
so that users may assess it as an outlier. Additionally, Pearson residuals
will be plotted as a histogram and a QQ-plot to check for deviances from model
assumptions. We assume that residuals will follow a normal distribution with
a mean of 0. Thus, we expect the histogram to follow a bell curve and the
QQ-plot to be plotted along the y=x line.

## General Usage

Mutation data should first be summarised by sample using `calculate_mf()`.
The `mf_data` should be output as a summary table. Be sure to retain the
columns for experimental variables of interest using the
`retain_metadata_cols` parameter.

You may specify factors and covariates for your model using the `fixed_effects`
and `random_effects` parameters respectively. If more than one fixed_effect is
supplied, then you may specify whether you wish to test the interaction between
your fixed_effects using the `test_interaction` parameter. 

You must specify the columns in your `mf_data` that contain the mutation counts
and the total_depth per sample using the `muts` and `total_counts` parameters
respectively. 

To perform pairwise comparisons between levels of your fixed effects, supply a
constrast table using the `contrasts` parameter. This can either be a data
frame  or a filepath to a text file that will be loaded into R. The table must
consist of two columns, each containing levels within the fixed_effects.
The level in the first column will be compared to the level in the second
column for each row. You should also provide the reference level for each
fixed effect using the `reference_level` parameter. If you specify
multiple pairwise comparisons, then the p-values will automatically be
corrected for multiple comparisons using the Sidak method. For multiple
fixed effects, the user must include levels for all fixed_effects in each
value of the contrasts table. Within each value, the levels of the different
fixed_effects should be seperated by a colon.

## Output

The function will output a list of results.

- model: the glmm or glmer model object.
- model_formula: the formula given to glm or glmer.
- model_data: the supplied `mf_data` with added column for Pearson residuals.
- summary: the summary of the model.
- anova: the analysis of variance for models with two or more fixed_effects.
See
[car::Anova.](https://www.rdocumentation.org/packages/car/versions/1.0-9/topics/Anova).
- residuals_histogram: the Pearson residuals plotted as a histogram. This is
used to check whether the variance is normally distributed. A symmetric
bell-shaped histogram, evenly distributed around zero indicates that the
normality assumption is likely to be true.
- residuals_qq_plot: the Pearson residuals plotted in a quantile-quantile plot.
For a normal distribution, we expect points to roughly follow the y=x line.  
- point_estimates_matrix: the contrast matrix used to generate point-estimates
for the fixed effects. 
- point_estimates: the point estimates for the fixed effects.
- pairwise_comparisons_matrix: the contrast matrix used to conduct the pairwise
comparisons specified in the `contrasts`.
- pairwise_comparisons: the results of pairwise comparisons specified in the
`contrasts`.

## Examples

### Example 1. Model by Dose {.tabset}

*Example 1. Model the effect of dose on MF. Our example data consists of 24*
*mouse samples, exposed to 3 doses of BaP or a vehicle control. Dose Groups*
*are : Control, Low, Medium, and High. We will determine if the MFmin of each*
*BaP dose group is significantly increased from that of the Control group.*
*Example data is the same as that used in the vignette. See the vignette*
*for information on importing and filtering the data.*

First, we will calculate MF per sample while retaining the dose column.

```{r calculate-mf-global}
# load example data:
example_data <- eh[["EH9861"]]

mf_data_global <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "sample",
  subtype_resolution = "none",
  retain_metadata_cols = c("dose_group", "dose")
)
```

Next, we will build a contrasts table that compares each BaP dose group
back to the control group.

```{r make-contrasts1}
# Create a contrasts table for pairwise comparisons
contrasts_table <- data.frame(
  col1 = c("Low", "Medium", "High"),
  col2 = c("Control", "Control", "Control")
)
```

Finally, we will run the model.

```{r model-mf-dose, fig.show='hide'}
# Run the model
model_by_dose <- model_mf(
  mf_data = mf_data_global,
  fixed_effects = "dose_group",
  muts = "sum_min",
  total_count = "group_depth",
  reference_level = "Control",
  contrasts = contrasts_table
)
```

#### Residuals Histogram

```{r hist-caption1, include=FALSE}
caption <- paste(
  "GLM residuals of MFmin modelled as an effect of Dose.",
  "x is pearson's residuals, y is frequency.",
  "Plotted to validate model assumptions. n = 24."
)
```

```{r model-dose-hist, echo=FALSE, fig.small=TRUE, fig.cap=caption}
model_data <- model_by_dose$model_data
hist_data <- hist(model_data$residuals, plot = FALSE)
ylim_max <- max(hist_data$counts) + 1
plot <- hist(model_data$residuals,
  main = "Pearson Residuals",
  col = "yellow",
  ylim = c(0, ylim_max)
)
```

#### Residuals QQ-plot

```{r qq-caption1, include=FALSE}
caption <- paste(
  "GLM residuals of MFmin modelled as an effect of Dose",
  "expressed as a quantile-quantile plot.",
  "Y is the pearson's residuals of the model in ascending order",
  "x is the quantiles of standard normal distribution for n of 24.",
  "Plotted to validate model assumptions."
)
```

```{r model-dose-qq, echo=FALSE, fig.small=TRUE, fig.cap=caption}
qqplot <- stats::qqnorm(model_data$residuals, main = "QQ Plot of Residuals")
stats::qqline(model_data$residuals, col = "red")
```

#### Model Summary
```{r model-dose-summary}
model_by_dose$summary
```

#### Model Estimates

**Table 18.** model_by_dose$point_estimates: Estimated Mean Mutation Frequency per Dose.

```{r model-dose-est, echo=FALSE}
DT::datatable(
  model_by_dose$point_estimates,
  options = list(
    columnDefs = list(
      list(targets = 1:4,
           render = DT::JS("function(data, type, row, meta) {
            return data.toExponential(2); 
          }"))
    ),
    rowCallback = DT::JS("function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }")
  )
)
```

#### Pairwise Comparisons

**Table 19.** model_by_dose$pairwise_comparisons

```{r model-dose-comp, echo=FALSE}
DT::datatable(
  model_by_dose$pairwise_comparisons,
  options = list(
    scrollX = TRUE,
    columnDefs = list(
      list(targets = c("p.value", "adj_p.value"),
        render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }")
      )
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
      $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
) %>%
  DT::formatRound(columns = c(1:3, 6:7), digits = 2)
```

### Example 2. Model by Dose and Target {.tabset}

*Example 2. Model the effects of dose and genomic locus on MF. Sequencing for*
*the example data was done on a panel of 20 genomic targets. We will determine*
*if the MF of each BaP dose group is significantly different from the Control*
*individually for all 20 targets. In this model, dose group and target label*
*will be our fixed effects. We include the interaction between the two fixed*
*effects. Because sample will be a repeated measure, we will use it as a random*
*effect. See the vignette for more information on how we imported and filtered*
*this dataset.*

First, calculate MF for each sample and genomic locus, while retaining the dose
column.

```{r mf-rg}
mf_data_rg <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = c("sample", "label"),
  subtype_resolution = "none",
  retain_metadata_cols = "dose_group"
)
```

Next, create a contrasts table the compares each dose group back to the
control for each genomic locus:.

```{r make-contrasts2}
combinations <- expand.grid(dose_group = unique(mf_data_rg$dose_group),
                            label = unique(mf_data_rg$label))
combinations <- combinations[combinations$dose_group != "Control", ]
combinations$col1 <- with(combinations, paste(dose_group, label, sep = ":"))
combinations$col2 <- with(combinations, paste("Control", label, sep = ":"))
contrasts2 <- combinations[, c("col1", "col2")]
```

Finally, run the model. As this model is more complex, we will improve
covergence by supplying the control argument, which is passed directly to
`lme4::glmer()`.

``` {r model-mf-rg, fig.show='hide'}
model_by_target <- model_mf(mf_data = mf_data_rg,
  fixed_effects = c("dose_group", "label"),
  test_interaction = TRUE,
  random_effects = "sample",
  muts = "sum_min",
  total_count = "group_depth",
  contrasts = contrasts2,
  reference_level = c("Control", "chr1"),
  control = lme4::glmerControl(optimizer = "bobyqa",
                               optCtrl = list(maxfun = 2e5))
)
```

#### Residuals Histogram

```{r hist-caption2, include=FALSE}
caption <- paste(
  "GLMM residuals of MFmin modelled as an effect of Dose and Genomic Target.",
  "x is pearson's residuals, y is frequency.",
  "Plotted to validate model assumptions. n = 24."
)
```

```{r model-rg-hist, echo=FALSE, fig.small=TRUE, fig.cap=caption}
model_data <- model_by_target$model_data
hist_data <- hist(model_data$residuals, plot = FALSE)
ylim_max <- max(hist_data$counts) + 1
plot <- hist(model_data$residuals,
  main = "Pearson Residuals",
  col = "yellow",
  ylim = c(0, ylim_max)
)
```

#### Residuals QQ-plot

```{r qq-caption2, include=FALSE}
caption <- paste(
  "GLMM residuals of MFmin modelled as an effect of Dose and Genomic Target",
  "expressed as a quantile-quantile plot.",
  "Y is the pearson's residuals of the model in ascending order",
  "x is the quantiles of standard normal distribution for n of 24.",
  "Plotted to validate model assumptions."
)
```

```{r model-rg-qq, echo=FALSE, fig.small=TRUE, fig.cap=caption}
qqplot <- stats::qqnorm(model_data$residuals, main = "QQ Plot of Residuals")
stats::qqline(model_data$residuals, col = "red")
```

#### Model Summary
```{r model-rg-summary}
model_by_target$summary
```

#### ANOVA

```{r model-rg-anova}
model_by_target$anova
```

#### Model Estimates

**Table 20.** model_by_target$point_estimates: Estimated Mean Mutation Frequency per Dose and Genomic Target.

```{r model-rg-est, echo=FALSE}
DT::datatable(
  model_by_target$point_estimates,
  options = list(
    columnDefs = list(
      list(targets = 1:4,
           render = DT::JS("function(data, type, row, meta) {
            return data.toExponential(2); 
          }"))
    ),
    rowCallback = DT::JS("function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }")
  )
)
```

#### Pairwise Comparisons

**Table 21.** model_by_target$pairwise_comparisons.

```{r model-rg-comp, echo=FALSE}
DT::datatable(
  model_by_target$pairwise_comparisons,
  options = list(
    scrollX = TRUE,
    columnDefs = list(
      list(targets = c("p.value", "adj_p.value"),
        render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }")
      )
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
      $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
) %>%
  DT::formatRound(columns = c(1:3, 6:7), digits = 2)
```

## Plot Model Results {.tabset}

Plot the results of `model_mf()` using `plot_model_mf()`. This function will
create a bar or line plot of the point estimates. Users can include the
estimated standard error as error bars with `plot_error_bars` and add
significance labels based on the pairwise comparisons with `plot_signif`.
The function can plot model results with up to two fixed effects. Users must
specify which fixed effect should be represented on the x-axis with `x_effect`.
The other fixed effect will be represented by the fill aesthetic.

When adding significance labels, the function will generate a unique symbol for
each unique level or combination of levels in column 2 of the contrasts table.
If using two fixed effects, but your comparisons only contrast levels of one
fixed effect, you may specify this using the `ref_effect` parameter. Supply it
with the fixed effect that is being contrasted to reduce the number of unique
symbols generated.

The output is a ggplot object that can be modified with ggplot2.


### Plot Model by Dose

```{r plot-model-cap, include=FALSE}
caption <- paste(
  "Mean Mutation Frequency Minimum (mutations/bp) per Dose",
  "estimated using a generalized linear model. Error bars are the S.E.M.",
  "Symbols indicate significance differences (p < 0.05)."
)
```

```{r plot-model-dose, fig.cap=caption}
plot <- plot_model_mf(
  model_by_dose,
  plot_type = "bar",
  x_effect = "dose",
  plot_error_bars = TRUE,
  plot_signif = TRUE,
  x_order = c("Control", "Low", "Medium", "High"),
  x_label = "Dose Group",
  y_label = "Estimated Mean Mutation Frequency (mutations/bp)"
)
plot
```

### Plot Model by Dose and Target

In this example, we only made comparisons between dose groups. For each
contrast, we held the label (target) constant. Thus, we will set the
ref_effect to dose_group so that significance labels are generated to
indicate differences in dose, and not label.

```{r plot-model-cap2, include=FALSE}
caption <- paste(
  "Mean Mutation Frequency Minimum (mutations/bp) per Genomic Target and Dose",
  "estimated using a generalized linear mixed model. Error bars are the SEM.",
  "Symbols indicate significance differences (p < 0.05) between dose levels",
  "for individual genomic regions."
)
```

```{r plot-model-rg, fig.wide=TRUE, fig.cap=caption}
# Define the order of the genomic targets for the x-axis:
# We will order them from lowest to highest MF at the High dose.
label_order <- model_by_target$point_estimates %>%
  dplyr::filter(dose_group == "High") %>%
  dplyr::arrange(Estimate) %>%
  dplyr::pull(label)

# Define the order of the doses for the fill
dose_order <- c("Control", "Low", "Medium", "High")

plot <- plot_model_mf(
  model = model_by_target,
  plot_type = "bar",
  x_effect = "label",
  plot_error_bars = TRUE,
  plot_signif = TRUE,
  ref_effect = "dose_group",
  x_order = label_order,
  fill_order = dose_order,
  x_label = "Target",
  y_label = "Mutation Frequency (mutations/bp)",
  fill_label = "Dose",
  plot_title = "",
  custom_palette = c("#ef476f",
                     "#ffd166",
                     "#06d6a0",
                     "#118ab2")
)
# Rotate the x-axis labels for clarity using ggplot2 functions.
plot <- plot + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90))
plot
```

## Exporting Model Results

`write_excel()` will extract the mf_data, point_estimates, and
pairwise_comparisons from `model_mf()` output to write to an excel workbook.
To export model results, provide `write_excel` with the output of `model_mf`
and set the `model_results` parameter to TRUE.

*Example 3. Export model results*
```{r write-excel-model, eval=FALSE}
write_excel(
  model_by_dose,
  workbook_name = "Example_model",
  model_results = TRUE
)
```

# Benchmark Dose Modeling

Dose-response models are essential for quatitative risk assessment of
mutagenicity, as they provide a framework to evaluate the levels at which
exposure to a substance might cause an adverse effect. The **benchmark dose**
(BMD) is a dose that produces a predetermined change in the measured response,
defined as the **benchmark response** (BMR). The BMD is used as a point of
departure to derive human health-based guidance values to inform regulatory
risk assessment such as the reference dose (RfD), the derived no-effect level
(DNEL) or the acceptable daily intake (ADI) [@efsa-2022].

The BMD is estimated by applying various mathmatical models to fit the
dose-response data. Some requirements must be met before modelling the BMD.
There must be a clear dose-response trend in the MF data. We suggest using
`model_mf()` to test for significant increases in MF with dose prior to running
a BMD analysis. In general, studies with more dose groups and a graded
monotonic response with dose will be more useful for BMD analysis. A minimum of
three dose groups + 1 control group is suggested. Datasets in which a response
is only observed at the high dose are usually not suitable for BMD modeling.
However, if the one elevated response is near the BMR, adequate BMD computation
may result. For a better estimate of the BMD, it is preferable to have studies
with one or more doses near the level of the BMR.

Protection and safety authorities recommend the use of model averaging to
determine the BMD and its confidence intervals. Model averaging incorporates
information across multiple models to acount for model uncertainty, allowing
the BMD to be more accurately estimated.

Ideally, the BMR would be based on a consensus scientific definition of what
minimal level of change in MF is biologically significant. Currently, the
default provided by this package calculates the BMD at a 50% relative increase
in MF from the background. This BMR was selected based on previous
recommendations for genotoxicity assessment by @white-2020.

MutSeqR provides `bmd_proast` for BMD modelling, which runs a modified version of the
[proast71.1](https://www.rivm.nl/en/proast) R library that is parametirized instead of
menu-based.

`bmd_proast` will analyze the continuous, individual MF data following a log
transformation. PROAST uses four families of nested models: exponential, Hill,
inverse exponential, and log-normal. The Akaike information criterion (AIC) is
used to select best fit. BMD confidence intervals are assessed by the Maximum
Likelihood Profile method, or by model averaging via bootstrapping (recommended).
BMR values are a user-defined relative increase in MF from the
control. Alternatively, users may set the BMR as one standard-deviation from
the control.

## General Usage: bmd_proast

Supply `bmd_proast` with per-sample mf data calculated using `calculate_mf()`,
with the dose column retained in the summary table. Dose column must be numeric.

Specify the column that contains the numeric dose values in `dose_col`. The
function can model more than one response variable at once. Supply all response
variables to `response_col`. If you wish to include a covariate in the analysis,
supply the covariate variable to `covariate_col`. PROAST will assess if the BMD
values differ significantly between levels of the covariate and give a BMD
estimate for each.

It is highly recommended to use model averaging when calculating the BMD
confidence intervals. Specify the number of bootstraps to run with
`num_bootstraps`. The recommended value is 200, but be aware that this may
take some time to run. PROAST model averaging will return the upper and lower
90% BMD confidence intervals. MutSeqR calculates the model-averaged BMD value
as the median BMD of all bootstrap runs.

Users may choose to generate model plots with `plot_results`. Plots may be
automatically saved to an output directory specified in `output_path`.
Alternatively, when output_path is NULL, plots will automcatically
be displayed and then returned within a list alongside the BMD
summary results.

The function will output a data frame of final results, including a BMD
estimate for each model family and the model averaging results, if applicable.
Users may access the raw, unparsed PROAST results by setting
`raw_results = TRUE`.

## Example

*Example 2.1. Calculate the BMD with model averaging for a 50% relative*
*increase in MF from control. This will be calculated for both MFmin and MFmax.*
*See the vignette to see information on how we imported and filtered the data.*

Fist, calculate MF per sample while retaining the numeric dose column:

```{r calculate-mf-global-bmd}
# load example data:
example_data <- eh[["EH9861"]]

mf_data_global <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "sample",
  subtype_resolution = "none",
  retain_metadata_cols = c("dose")
)
```

Next, run bmd_proast()

```{r proast, results='hide', fig.show = 'hide'}
proast_results <- bmd_proast(
  mf_data = mf_data_global,
  dose_col = "dose",
  response_col = c("mf_min", "mf_max"),
  bmr = 0.5,
  model_averaging = TRUE,
  num_bootstraps = 10, # recommended value 200
  plot_results = TRUE,
  output_path = NULL
)
```

## BMD Results {.tabset}

### Summary

**Table 22.** BMD Estimates by PROAST.

```{r proast-results, echo=FALSE}
DT::datatable(proast_results$summary) %>%
  DT::formatRound(columns = 3:8, digits = 2)
```

### Exponential & Hill Models

```{r bmd-exphill-cap, include=FALSE}
caption1 <- paste(
  "The fitted curve of the selected Exponenetial (left) and Hill (right)",
  "models. Data is MFmin by Dose following",
  "log-transformation. Individual data points are plotted using small",
  "triangles. The geometric mean at each dose is plotted",
  "as a large triangle. The estimated BMD is indicated by a dotted line."
)
caption2 <- paste(
  "The fitted curve of the selected Exponenetial (left) and Hill (right)",
  "models. Data is MFmax by Dose following",
  "log-transformation. Individual data points are plotted using small",
  "triangles. The geometric mean at each dose is plotted",
  "as a large triangle. The estimated BMD is indicated by a dotted line."
)
```

```{r bmd-exphill-plot1, fig.cap=caption1}
replayPlot(proast_results$mf_min_Expon_HILL)
```

```{r bmd-exphill-plot2, fig.cap=caption2}
replayPlot(proast_results$mf_max_Expon_HILL)
```

###  Inverse Exponential & Log-Normal Models

```{r bmd-invln-cap, include=FALSE}
caption1 <- paste(
  "The fitted curve of the selected Inverse-Exponenetial (left) and",
  "Log-Normal (right) models. Data is MFmin by Dose",
  "following log-transformation. Individual data points are plotted using",
  "small triangles. The geometric mean at each dose",
  "is plotted as a large triangle. The estimated BMD is indicated by a dotted",
  "line."
)
caption2 <- paste(
  "The fitted curve of the selected Inverse-Exponenetial (left) and",
  "Log-Normal (right) models. Data is MFmax by Dose",
  "following log-transformation. Individual data points are plotted using",
  "small triangles. The geometric mean at each dose",
  "is plotted as a large triangle. The estimated BMD is indicated by a dotted",
  "line."
)
```

```{r bmd-invln-plot1, fig.cap=caption1}
replayPlot(proast_results$mf_min_InvExp_LN)
```

```{r bmd-invln-plot2, fig.cap=caption2}
replayPlot(proast_results$mf_max_InvExp_LN)
```

###  Bootstrap Curves

```{r bmd-boot-cap, include=FALSE}
caption1 <- paste(
  "The bootstrap curves based on model averaging. Data is MFmin (left) and MFmax (right) by",
  " Dose with log-transformation. The geometric mean at each dose",
  "is plotted as a triangle. Number of bootstraps = 10 (although the recommended value is 200)."
)
```

```{r bmd-boot-plot2, fig.cap=caption1}
replayPlot(proast_results$mf_max_bootstrap_curves)
```

### Cleveland Plots

```{r bmd-cvd-cap, include=FALSE}
caption1 <- paste(
  "MFmin. The BMD estimate for each model plotted as a red point alongside the 90%",
  "confidence intervals. The size of the BMD point represents the model weight",
  "assigned during model averaging based on AIC."
)
caption2 <- paste(
  "MFmax. The BMD estimate for each model plotted as a red point alongside the 90%",
  "confidence intervals. The size of the BMD point represents the model weight",
  "assigned during model averaging based on AIC."
)
```

```{r bmd-cvd-plot1, fig.cap=caption1}
proast_results$mf_min_cleveland
```

```{r bmd-cvd-plot2, fig.cap=caption2}
proast_results$mf_max_cleveland
```

## Plot BMD CIs

`plot_ci()` creates a confidence interval (CI) plot of BMD results for easy
comparison of BMDs between response variables.

The BMD values can be plotted on the log scale, if required.

*Example 2.2. Compare the estimated BMD for MFmin (50% relative increase)*
*by PROAST versus ToxicR (another BMD software available on Github,*
* calculated independently)*

```{r ci-plot-caption, include=FALSE}
caption <- paste(
  "Benchmark dose with 90% confidence intervals representing the dose at,",
  "which a 50% increase in mutation frequency occurs from reference level.",
  "Calculated using ROAST software. Black points represent the BMD,",
  "red points the BMDL, and blue points, the BMDU"
)
```

```{r plot-ci, fig.cap=caption}
plot_results <- data.frame(
  Response = c("PROAST", "ToxicR"),
  BMD = c(9.111, 9.641894),
  BMDL = c(7.38, 8.032936),
  BMDU = c(10.9, 10.97636)
)
plot <- plot_ci(
  data = plot_results,
  order = "asc",
  x_lab = "Dose (mg/kg-bw/d)",
  y_lab = "BMD Method"
)
plot
```

# References

# Appendix

## Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```
