---
title: "MutSeqR: Other Features"
author:
- name: Annette E. Dodge
  affiliation: Environmental Health Science and Research Bureau, Health Canada, Ottawa, ON, Canada.
- name: Matthew J. Meier
  affilitation: Environmental Health Science and Research Bureau, Health Canada, Ottawa, ON, Canada.
  email: matthew.meier@hc-sc.gc.ca
package: MutSeqR
output:
  BiocStyle::html_document:
    toc: true
    toc_float: yes
    code_folding: show
    code_download: yes
    css: custom.css
---
  
```{r setup, include=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(DT)
library(htmltools)
```
# Load MutSeqR and Example Data

```{r load-lib}
library(MutSeqR)
```

```{r load-ex}
library(ExperimentHub)
# load the index
eh <- ExperimentHub()
```

# Bubble Plots

`plot_bubbles` is used to visually represent the distribution and density of
recurrent mutations. Each mutation is in a given group is represented by a
bubble whose size is scaled on either the `alt_depth` or the `vaf`. Thus a
highly reccurent mutation is represented by a large bubble. These plots make
it easy to determine if MFmax is driven by a few highly recurrent mutations
versus serveral moderately recurrent mutations.

Plots can be facetted by user-defined groups, and bubbles can be coloured by
any variable of interest to help discern patterns in mutation recurrence.

*Example 1. Plot mutations per dose group, bubbles coloured by base-6 subtype*

```{r bubbles-cap, include=FALSE}
caption <- paste(
  "Multiplet mutations plotted per Dose. Each circle represents a mutation,",
  "coloured by mutation subtype. The size of the circle is scaled by the",
  "mutation's alternative depth."
)
```

```{r bubble-plot, fig.wide=TRUE, fig.cap=caption}
# load example data:
example_data <- eh[["EH9861"]]

plot <- plot_bubbles(
  mutation_data = example_data,
  size_by = "alt_depth",
  facet_col = "dose_group",
  color_by = "normalized_subtype"
)
plot
```

# Retrieve Sequences

`get_seq()` will retrive raw nucleotide sequences for specified genomic
intervals. The function can retrieve sequences in one of two ways. Either
specify an installed BSgenome from which to retrieve sequences or retrieve
sequences through API request to the UCSC database based on supplied species
and genome assembly.

Supply regions with a filepath, data frame or GRanges object containing the
specified genomic intervals. TwinStrand's Mutagenesis Panels are stored in
package files and can easily be retrieved. 

Supply the function with the appropriate BS_genome from which to retrieve
sequences. You can use the find_BS_genome() function to identify a BS genome
to use based on species and genome assembly. Be sure to install the BS genome
prior to running get_seq() with BiocManager().

To retrieve sequences from the UCSC data base, please input the species and
assembly genome to the species and genome parameters.

**BS_genome, species, or genome*
*do not need to be specified for the Mutagenesis Panels as this information*
*is stored*.

Sequences are returned within a *GRanges* object.

*Example 2.1. Retrieve the sequences for our example's target panel,*
*TwinStrand's Mouse Mutagenesis Panel*
```{r get-seq}
regions_seq <- get_seq(regions = "TSpanel_mouse")
regions_seq
```

*Example 2.2. Retrieve sequences for a custom interval of regions. We will use*
*the Human Mutagenesis Panel as an example.*
```{r get-seq-custom}
# We will load the TSpanel_human regions file to obtain
# an example list of regions.
human <- load_regions_file("TSpanel_human")
regions_seq <- get_seq(regions = human,
                       is_0_based_rg = FALSE,
                       BS_genome = find_BS_genome("human", "hg38"),
                       padding = 0)
regions_seq
```

*Example 2.3. Retrieve sequences using UCSC database.*
```{r get-seq-ucsc}
regions_seq <- get_seq(regions = human,
                       is_0_based_rg = FALSE,
                       padding = 0,
                       species = "human",
                       genome = "hg38",
                       ucsc = TRUE)
```

Sequences can be exported as FASTA files with `write_reference_fasta()`.
Supply this function with the GRanges object containing the sequences of the
regions. Each one will be written to a single FASTA file. The FASTA file will
be saved to the `output_path`. If NULL, the file will be saved to the working
directory.

```{r write-ref-fasta, eval=FALSE}
write_reference_fasta(regions_seq, output_path = NULL)
```

# References

# Appendix

## Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```
