---
title: "MutSeqR: Error-Corrected Sequencing (ECS) Analysis For Mutagenicity Assessment"
author:
- name: Annette E. Dodge
  affiliation: Environmental Health Science and Research Bureau, Health Canada, Ottawa, ON, Canada.
- name: Matthew J. Meier
  affilitation: Environmental Health Science and Research Bureau, Health Canada, Ottawa, ON, Canada.
  email: matthew.meier@hc-sc.gc.ca
package: MutSeqR
output:
  BiocStyle::html_document:
    toc: true
    toc_float: yes
    code_folding: show
    code_download: yes
    css: custom.css
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{MutSeqR: Error-Corrected Sequencing (ECS) Analysis For Mutagenicity Assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(DT)
library(htmltools)
```

# Introduction

In this module we will provide instructions on how to import, filter, and
summarize mutation data generated by Error-correct next-generation sequencing (ECS).
For information on how to use MutSeqR for statistical analysis of the data, see
the relevant modules:

- Modelling Mutation Frequencies (Generalized Linear Modelling and BMD)
- Analyzing Mutation Spectra (Comparison of Spectra and Mutation Signature Analysis)
- Target Analysis
- Other Features (visualization of clonal mutations, retrieve target sequences)

## What is ECS?

Error-corrected next-generation sequencing (ECS) uses various methods to
combine multiple independent raw sequence reads derived from an original
starting molecule, thereby subtracting out artifacts introduced during
sequencing or library preparation. This results in a highly accurate
representation of the original molecule. ECS is particularly useful for
detecting rare somatic mutations (or mutations induced in germ cells),
such as those that arise from mutagen exposure or other sources of DNA
damage. ECS is a powerful tool for assessing the mutagenicity of chemicals,
drugs, or other agents, and can be used to identify the mutational signatures
of these agents. ECS can also be used to detect rare mutations in cancer or
other diseases, and to track the clonal evolution of these diseases over time.

For more background on how ECS works and its context in regulatory toxicology
testing and genetic toxicology, see the following articles:

- @menon-2023
- @marchetti-2023a
- @marchetti-2023b
- @kennedy-2014

This R package is meant to facilitate the import, cleaning, and analysis of ECS
data, beginning with a table of variant calls or a variant call file (VCF). The
package is designed to be flexible and enable users to perform common
statistical analyses and visualisations.

# Installation
Install the package from github:

```{r install-git, eval = FALSE}
install.packages("devtools")

devtools::install_github(
  "EHSRB-BSRSE-Bioinformatics/MutSeqR"
)
```

Install the package from Bioconductor (currently not implemented):

```{r install-bioc, eval=FALSE}
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("MutSeqR")
```

Load the package

```{r load-lib}
library(MutSeqR)
```

# Data import

The main goal of MutSeqR is to generate summary statistics, visualizations,
exploratory analyses, and other post-processing tasks such as mutational
signature analysis or generalized linear modeling. Mutation data should be
supplied as a table of variants with their genomic positions. Mutation data
can be imported as either VCF files or as tabular data using the functions
`import_vcf_data` and `import_mut_data`, respectively. It is reccomended that
files include a record for every sequenced position, regardless of whether a
variant was called or not, along with the `total_depth` for each record. This
enables site-specific depth calculations that are required for the calculation
of mutation subtype frequencies ad other site-specific frequemcies. The data
set can be pared down later to include only mutations of interest (SNVs,
indels, SVs, or any combination). 

Table: (\#tab:required-columns) Required columns for mutation data import.

| **Column** | **VCF Specification** | **Definition** |
|------------|-----------------------|----------------|
| contig | CHROM | The name of the reference sequence. |
| start | POS | The start position of the feature. 0-based coordinates are accepted but will be changed to 1-based during import. |
| end | INFO(END) | The half-open end position of the feature in contig. |
| ref | REF | The reference allele at this position. |
| alt | ALT | The left-aligned, normalized, alternate allele at this position. |
| sample | INFO(sample) or Genotype Header | A unique identifier for the sample library. For VCF files, this field may be provided in either the INFO field, or as the header to the GENOTYPE field. |
| *SUGGESTED FIELDS* | | |
| alt_depth | VD | The read depth supporting the alternate allele. If not included, the function will assume an alt_depth of 1 at variant sites. |
| total_depth or depth | AD or DP | The total read depth at this position. This column can be “total_depth” which excludes N-calls, or “depth”, which includes N-calls, if “total_depth” is not available. For VCF files, the total_depth is calculated as the sum of AD. DP is equivalent to "depth". |

VCF files should follow the VCF specification (version 4.5; @danecek-2011).
VCF files may be bg/g-zipped. Each individual VCF file should contain the
mutation data for a single sample library. Multiple alt alleles called for a
single position should be represented as separate rows in the data. All extra
columns, INFO fields, and FORMAT fields will be retained upon import.

Upon import, records are categorized within the `variation_type` column based
on their REF and ALT. Categories are listed below.

Table: (\#tab:variation-types) Definitions of Variation types.

| variation_type | Definition |
|------------------|------------|
| no_variant | No variation, the null-case. |
| snv | Single nucleotide variant. |
| mnv | Multiple nucleotide variant. |
| insertion | Insertion. |
| deletion | Deletion. |
| complex | REF and ALT are of different lengths and nucleotide compositions. |
| symbolic | Structural variant |
| ambiguous | ALT contains IUPAC ambiguity codes. |
| uncategorized | The record does not fall into any of the preceding categories. |

Additional columns are created to further characterise variants.

Table: (\#tab:mut-columns) Definitions of Mutation data columns.

| Column Name        | Definition                                        |
|--------------------|---------------------------------------------------|
|  short_ref | The reference base at the start position. |
| normalized_ref | The short_ref in C/T (pyrimidine) notation for this position. Ex. `A` -> `T`, `G` -> `C` |
| context | The trinucleotide context at this position. Consists of the reference base and the two flanking bases. Sequences are retrieved from the appropriate BS genome. Ex. `TAC` |
| normalized_context | The trinucleotide context in C/T (pyrimidine) notation for this position (Ex. `TAG` -> `CTA`) |
| variation_type | The type of variant (no_variant, snv, mnv, insertion, deletion, complex, sv, ambiguous, uncategorized) |
| subtype | The substitution type of the snv variant (12-base spectrum; Ex. `A>C`) |
| normalized_subtype | The snv subtype in C/T (pyrimidine) notation (6-base spectrum; Ex. `A>C` -> `T>G`) |
| context_with_mutation | The snv subtype including the two flanking nucleotides (192-base spectrum; Ex. `T[A>C]G`) |
| normalized_context_with_mutation | The snv subtype in C/T (pyrimidine) notation including the two flanking nucleotides (96-base spectrum; Ex. `T[A>C]G` -> `C[T>G]A`) |
| nchar_ref | The length (in bp) of the reference allele. |
| nchar_alt | The length (in bp) of the alternate allele. |
| varlen | The length (in bp) of the variant. |
| ref_depth | The depth of the reference allele. Calculated as `total_depth` - `alt_depth`, if applicable. |
| vaf | The variant allele fraction. Calculated as `alt_depth`/`total_depth` |
| gc_content | % GC of the trinucleotide context at this position. |
| is_known | A logical value indicating if the record is a known variant; i.e. ID field is not NULL. |
| row_has_duplicate | A logical value that  flags rows whose position is the same as that of at least one other row for the same sample. |

## General Usage

Indicate the filepath to your mutation data file(s) using the `vcf_file`/
`mut_file` parameter. This can be either a single file or a directory
containing multiple files. Should you provide a directory, then all files
within will be bound into a single data frame.

An important component of importing your data for proper use is to assign each
mutation to a biological sample, and also make sure that some additional
information about each sample is present (ex., a chemical treatment, a dose,
etc.). This is done by providing `sample_data`. This parameter can take a data
frame, or it can read in a file if provided it with a filepath. If using a
filepath, specify the proper delimiter using the `sd_sep` parameter. Sample
metadata will be joined with the mutation data using the "sample" column to
capture that information and associate it with each mutation.

Specify the appropriate BS genome with which to populate the context column.
Prior to data import, ensure that the BS genome is installed. Supply the pkgname
to the BS_genome parameter in the import function. If you are not sure which
BS genome you should use, you can use the find_BS_genome() function. This function
will browse [BSgenome::available.genomes](https://www.rdocumentation.org/packages/BSgenome/versions/1.40.1/topics/available.genomes)
for given organisms and genome assembly versions. Once you've identified the
appropriate BS genome, it can be installed using BiocManager().
Context information will be extracted from the BS genome.
Note, BSgenome offers genomes with masked sequences. If you wish to use the masked
version of the genome, you can find the right one with find_BS_genome(masked = TRUE).

## Examples {.tabset}

We provide an example data set taken from @leblanc-2022. This data
consists of 24 mouse bone marrow samples sequenced with Duplex Sequencing
using Twinstrand's Mouse Mutagenesis Panel of twenty 2.4kb targeted genomic
loci. Mice were exposed to three doses of benzo[a]pyrene (BaP) alongside
vehicle controls, n = 6. Example data is retrieved from MutSeqRData, an
ExperimentHub data package available via Bioconductor. Example data
can be retrieved from the ExperimentHub index (eh) through specific
accessors.

```{r load-MutSeqRData}
library(ExperimentHub)
# load the index
eh <- ExperimentHub()
# Query the index for MutSeqRData records and their associated accessors.
query(eh, "MutSeqRData")
```

Identify the BS genome: Sequencing data was aligned to the mm10 mouse genome.
```{r find_BS_genome}
mouse_BS_genome <- find_BS_genome(organism = "mouse", genome = "mm10")
print(mouse_BS_genome)
```

Install the appropriate BS genome:
```{r install_BS_genome, eval=FALSE}
BiocManager("BSgenome.Mmusculus.UCSC.mm10")
```

### VCF

*Example 1.1. Import the example .vcf.bgz file "Example import_vcf_data".*
*Provided is the genomic vcf.gz file for sample dna00996.1. It is comprised*
*of a record for all 48K positions sequenced for the Mouse Mutagenesis Panel*
*with the alt_depth and the tota_depth values for each record.*

```{r import-vcf}
example_file <- eh[["EH9859"]]

sample_metadata <- data.frame(
  sample = "dna00996.1",
  dose = "50",
  dose_group = "High"
)
# Import the data
imported_example_data <- import_vcf_data(
  vcf_file = example_file,
  sample_data = sample_metadata,
  BS_genome = mouse_BS_genome
)
```

**Data Table 1.** Mutation data imported from VCF for sample dna00996.1. Displays the first 6 rows.

```{r import-vcf-table, echo=FALSE}
DT::datatable(head(imported_example_data), options = list(scrollX = TRUE))
```

### Tabular

*Example 1.2. Import example tabular data "Example import_mut_data". This is*
*the equivalent file to the example vcf file. It is stored as an .rds file.*
*We will load the dataframe and supply it to `import_mut_data`. The mut_file*
*parameter can accept file paths or data frames as input.*

```{r import-mut}
example_data <- eh[["EH9857"]]

sample_metadata <- data.frame(
  sample = "dna00996.1",
  dose = "50",
  dose_group = "High"
)
# Import the data
imported_example_data <- import_mut_data(
  mut_file = example_data,
  sample_data = sample_metadata,
  BS_genome = find_BS_genome("mouse", "mm10"), # Must be installed!!
  is_0_based_mut = TRUE # indicates that the genomic coordinates are 0-based.
# Coordinates will be changed to 1-based upon import.
)
```

**Data Table 2.** Mutation data imported from a data.frame object for sample dna00996.1. Displays the first 6 rows.

```{r import-mut-table, echo=FALSE}
DT::datatable(head(imported_example_data), options = list(scrollX = TRUE))
```

## Region Metadata

Similar to sample metadata, you may supply a file containing the metadata of
genomic regions to the `regions` parameter. In the file, the genomic regions are
defined by the name of their reference sequence plus their start and end
coordinates. Additional columns are added to describe features of the genomic
regions (Ex. transcription status, GC content, etc.). Region metadata will be
joined with mutation data by checking for overlap between the target region
ranges and the position of the record.
The `regions` parameter can be either a filepath, a data frame, or a GRanges
object (see `r Biocpkg("GenomicRanges")`). Filepaths will be read using the
`rg_sep`.  Users can also choose from the built-in TwinStrand DuplexSeq™
Mutagenesis Panels by inputting "TSpanel_human",  "TSpanel_mouse", or
"TSpanel_rat". Required columns for the regions file are "contig", "start",
and "end". In a GRanges object, the required columns are "seqnames", "start",
and "end". Users must indicate whether the region coordinates are 0-based or
1-based with `is_0_based_rg`. Mutation data and region coordinates will be
converted to 1-based. If you do not wish to specify regions, then set the
`regions` parameter to NULL (default).

*Example 1.3. Add the metadata for TwinStrand's Mouse Mutagenesis panel to our*
*example tabular file.*

```{r import-mut-rg}
imported_example_data <- import_mut_data(
  mut_file = example_data,
  sample_data = sample_metadata,
  BS_genome = find_BS_genome("mouse", "mm10"),
  is_0_based_mut = TRUE,
  regions = "TSpanel_mouse"
)
```                                         

**Data Table 3.** Mutation data imported from a data.frame object for sample
dna00996.1 with added metadata for the Mouse Mutagenesis Panels sequencing
targets. Displays the first 6 rows. Added columns are 'target_size',
'label', 'genic_context', 'region_GC_content', 'genome', and 'in_regions'.

```{r import-mut-table-rg, echo=FALSE}
DT::datatable(head(imported_example_data), options = list(scrollX = TRUE))
```

Users can load the example TwinStrand Mutagenesis Panels with load_regions().
This function will output a GRanges object. Custom panels may also be loaded
with this function by providing their file path to the regions parameter.

```{r load-rg}
region_example <- load_regions_file("TSpanel_mouse")
region_example
```

## Custom Column Names

We recognize that column names may differ from those specified in
Table \@ref(tab:required-columns). Therefore, we have implemented some default
column name synonyms. If your column name matches one of our listed synonyms,
it will automatically be changed to match our set values. For example, your
`contig` column may be named `chr` or `chromosome`. After importing your data,
this synonymous column name will be changed to `contig`. Column names are
case-insensitive. A list of column name synonyms are listed below.

Table: (\#tab:name-sym) Predefined column name synonyms. Synonyms will be automatically changed to the default Column value upon import.

| Column | Synonyms |
|--------|----------|
| alt | alternate |
| alt_depth | alt_read_depth, var_depth, variant_depth, vd |
| context | sequence_context, trinucleotide_context, flanking_sequence |
| contig | chr, chromosome, seqnames |
| depth | dp |
| end | end_position |
| no_calls | n_calls, n_depth, no_depth |
| ref | reference, ref_value |
| sample | sample_name, sample_id |
| start | pos, position |
| total_depth | informative_somatic_depth |

If your data contains a column that is synonymous to one of the required
columns, but the name is not included in our synonyms list, your column
name may be substituted using the `custom_column_names` parameter. Provide
this parameter with a list of names to specify the meaning of column headers.

```{r import-mut-custom-cols, message=TRUE}
mut_data <- eh[["EH9858"]]

imported_example_data_custom <- import_mut_data(
  mut_file = mut_data,
  custom_column_names = list(my_contig_name = "contig",
                             my_sample_name = "sample")
)
```

**Data Table 4.** Mutation data imported from tabular file with custom column
names. Displays the first 6 rows. Custom column names are changed to default
values.

```{r import-mut-cust-table, echo=FALSE}
DT::datatable(head(imported_example_data_custom), options = list(scrollX = TRUE))
```

## Output

Mutation data can be output as either a data frame or a *GRanges* object
(see `r Biocpkg("GenomicRanges")`) for downstream analysis. Use the `output_granges`
parameter to specify the output. *GRanges* may faciliate use in other packages
and makes doing genomic based analyses on the ranges significantly easier. Most
downstream analyses provided by MutSeqR will use a data frame.

# Variant Filtering

Following data import, the mutation data may be filtered using the
`filter_mut()` function. This function will flag variants based on various
parameters in the `filter_mut` column. Variants flagged as TRUE in the
`filter_mut` column will be automatically excluded from downstream analyses
such as `calculate_mf()`, `plot_bubbles`, and `signature_fitting`. When
specified, this function may also remove records from the mutation data.

Flagging the variants in the `filter_mut` column will not remove them from
the mutation data, however, they will be excluded from mutation counts in
all downstream analyses. Filtered variants are retained in the data so that
their `total_depth` values may still be used in frequency calculations.

By default, all filtering parameters are disabled. Users should be mindful of
the filters that they use, ensuring first that they are applicable to their
data.

## Germline Variants

Germline variants may be identified and flagged for filtering by setting the
`vaf_cutoff` parameter. The variant allele fraction (vaf) is the fraction of
haploid genomes in the original sample that harbor a specific mutation at a
specific base-pair coordinate of the reference genome. Specifically, is it
calculated by dividing the number of variant reads by the total sequencing
depth at a specific base pair coordinate (`alt_depth` / `total_depth`). In
a typical diploid cell, a homozygous germline variant will appear on both
alleles, in every cell. As such, we expect this variant to occur on every
read, giving us a vaf = 1. A heterozygous germline variant occurs on one of
the two alleles in every cell, as such we expect this variant to occur on
about half of the reads, giving a vaf = 0.5. Somatic variants occur in only
a small portion of the cells, thus we expect them to appear in only a small
percentage of the reads. Typical vaf values for somatic variants are less
than 0.01 - 0.1. Setting the `vaf_cutoff` parameter will flag all variants that
have a vaf **greater than this value as germline** within the `is_germline`
column. It will also flag these variants in the `filter_mut` so as to exclude
them from downstream analyses.

vaf germline filtering is only applicable to users whose data was sequenced to
a sufficient depth. High coverage/low depth sequencing cannot be used to
calculate the vaf, thus it is recommended to filter germline mutations by
contrasting against a database of known polymorphisms, or by using conventional
whole-genome sequencing to identify germline variants for each sample. 

## Quality Assurance

The `filter_mut()` function offers some filtering options to ensure the quality
of the mutation data.

`snv_in_germ_mnv` = TRUE will flag all snvs that overlap with germline mnvs.
Germline mnvs are defined as having a vaf > vaf_cutoff. These snvs are often
artifacts from variant calling. No-calls in reads supporting the germline mnv
will create false minor-haplotypes from the original mnv that can appear as
sub-clonal snvs, thus such variants are excluded from downstream analyses.

`rm_abnormal_vaf` = TRUE This parameter identifies rows with abnormal vaf
values and **removes** them from the mutation data. Abnormal vaf values are
defined as being between 0.05-0.45 or between 0.55-0.95. In a typical diploid
organism, we expect variants to have a vaf ~0, 0.5, or 1, reflecting rare
somatic variants, heterozygous variants, and homozygous variants respectively.
Users should be aware of the ploidy of their model system when using this
filter. Non-diploid organisms may exhibit different vafs.

`rm_filtered_mut_from_depth = TRUE` Variants flagged in the `filter_mut`
column will have their `alt_depth` subtracted from the `total_depth`. When
TRUE, this parameter treats flagged variants as No-calls. This does not apply
to variants that were idenfied as germline variants.

## Custom Filtering

Users may use the `filter_mut()` function to flag or remove variants based on
their own custom column. Any record that contains the `custom_filter_val` value
within the `custom_filter_col` column of the mutation data will be either
flagged in the `filter_mut` column or, if specified by the `custom_filter_rm`
parameter, removed from the mutation data.

## Filtering by Regions

Users may remove rows that are either within or outside of specified genomic
regions. Provide the region ranges to the `regions` parameter. This may be
provided as either a filepath, data frame, or a GRanges object. `regions` must
contain `contig` (or `seqnames` for GRanges), `start`, and `end`.  The function
will check whether each record falls within the given regions. Users can define
how this filter should be used with `regions_filter`.
`region_filter = "remove_within"` will remove all rows whose positions overlap
with the provided regions. `region_filter = "keep_within"` will remove all rows
whose positions are outside of the provided regions. By default, records that are
> 1bp must start and end within the regions to count as being within the region.
`allow_half_overlap = TRUE` allow records that only start or end within the regions
but extend outside of them to be counted as being within the region. Twinstrand
Mutagenesis Panels may be used by setting `regions` to one of "TSpanel_human",
"TSpanel_mouse", or "TSpanel_rat".

## Output

The function will return the mutation data as a data.frame object with rows
removed or flagged depending on input parameters. Added columns are
'filter_mut', 'filter_reason', 'is_germline', and 'snv_mnv_overlaps'.
If `return_filtered_rows = TRUE` The function will return both the data.frame
containing the processed mutation data and a data.frame containing the rows
that were removed/flagged during the filtering process. The two dataframes
will be returned inside of a list, with names `mutation_data` and `filtered_rows`.
The function will also print the number of mutations/rows that were filtered
according to each parameter.

## Example

*Example 2. The example file "Example mutation data" is the output of*
*import_mut_data() run on all 24 mouse libraries from @leblanc-2022.*

*Filters used:*

- Filter germline variants: vaf < 0.01
- Filter snvs overlapping with germline variants and have their `alt_depth`
removed from their `total_depth`.
- Remove records outside of the TwinStrand Mouse Mutagenesis Panel.
- Filter variants that contain "EndRepairFillInArtifact" in the "filter"
column. Their `alt_depth` will be removed from their `total_depth`.

```{r filter-mut, message=TRUE}
# load the example data
example_data <- eh[["EH9860"]]

# Filter
filtered_example_mutation_data <- filter_mut(
  mutation_data = example_data,
  vaf_cutoff = 0.01,
  regions = "TSpanel_mouse",
  regions_filter = "keep_within",
  custom_filter_col = "filter",
  custom_filter_val = "EndRepairFillInArtifact",
  custom_filter_rm = FALSE,
  snv_in_germ_mnv = TRUE,
  rm_filtered_mut_from_depth = TRUE,
  return_filtered_rows = FALSE
)
```

**Data Table 5.** Filtered Mutation data for all 24 samples. Displayed are 6
variants that were flagged during filtering.

```{r filter-mut-table, echo=FALSE}
filter_table <- filtered_example_mutation_data %>%
  dplyr::filter(filter_mut == TRUE)
DT::datatable(head(filter_table), options = list(scrollX = TRUE))
```

# Calculating MF

Mutation Frequency (MF) is calculated by dividing the sum of mutations by the
sum of the `total_depth` for a given group (units mutations/bp). The function
`calculate_mf()` sums mutation counts and `total_depth` across user-defined
groupings within the mutation data to calculate the MF. Mutations can be
summarised across samples, experimental groups, and mutation subtypes for
later statistical analyses. Variants flagged as `TRUE` in the `filter_mut`
column will be excluded from the mutation sums; however, the *total_depth*
of these variants will be counted in the *total_depth* sum.

If the total_depth is not available, the function will sum the mutations across
groups, without calculating the mutation frequencies.

## Mutation Counting

Mutations are counted based on two opposing assumptions [@dodge-2023]:

The **Minimum Independent Mutation Counting Method** (min) Each mutation is
counted once, regardless of the number of reads that contain the non-reference
allele. This method assumes that multiple instances of the same  mutation
within a sample are the result of clonal expansion of a single mutational
event. When summing mutations across groups using the Min method (`sum_min`),
the `alt_depth` of each variant is set to 1. *Ex. For 3 variants with*
*`alt_depth` values of 1, 2, and 10, the `sum_min `= 3.*

The **Maximum Independent Mutation Counting Method** (max) Multiple identical
mutations at the same position within a sample are counted as independent
mutation events. When summing mutations across groups using the Max method
(`sum_max`), the `alt_depth` of each variant is summed unchanged. *Ex. for 3*
*variants with `alt_depth` values of 1, 2, and 10, the `sum_max` = 13.*

The Min and Max mutation counting methods undercount and overcount the
mutations, respectively. We expect some recurrent mutations to be a result of
clonal expansion. We also expect some recurrent mutations to arise
independently of each other. As we expect the true number of independent
mutations to be somewhere in the middle of the two counting methods, we
calculate frequencies for both methods. However, the Min mutation counting
method is generally recommended for statistical analyses to ensure
independance of values and because the Max method tends to increase the
sample variance by a significant degree. 

## Grouping Mutations

Mutation counts and `total_depth` are summed across groups that can be
designated using the `cols_to_group` parameter. This parameter can be set to
one or more columns in the mutation data that represent experimental variables
of interest. 

*Example 3.1. Calculate mutation sums and frequencies per sample.*
*The file "Example mutation data filtered" is the output of filter_mut()*
*from Example 2*

```{r mf-global}
# load example data:
example_data <- eh[["EH9861"]]

mf_data_global <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "sample",
  subtype_resolution = "none",
  retain_metadata_cols = c("dose_group", "dose")
)
```

**Table 6.** Global Mutation Frequency per Sample

```{r mf-global-table, echo=FALSE}
#| tbl-cap: "datatable {#tbl-datatable}"
DT::datatable(mf_data_global,
  options = list(
    columnDefs = list(
      list(targets = c("mf_min", "mf_max"),
           render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }"))
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
)
```

Alternatively, you can sum mutations by experimental groups such as `label`
(the genomic target). Counts and frequencies will be returned for every level
of the designated groups.

*Example 3.2. Calculate mutation sums and frequencies per sample and genomic target.*

```{r mf-rg}
mf_data_rg <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = c("sample", "label"),
  subtype_resolution = "none",
  retain_metadata_cols = "dose_group"
)
```

**Table 7.** Mutation Frequency per Sample and genomic target.

```{r mf-rg-table, echo=FALSE}
DT::datatable(mf_data_rg,
  options = list(
    columnDefs = list(
      list(targets = c("mf_min", "mf_max"),
           render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }"))
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
)
```

**calculate_mf() does not calculate the mean MF for any given group** If you
want to calculate the mean MF for a given experimental variable, you may group
by "sample" and retain the experimental variable in the summary table for
averaging using the `retain_metadata_cols` parameter. 

*Example 3.3. Calculate the mean MF per dose* 

```{r mean-mf}
mf_data_global <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "sample",
  subtype_resolution = "none",
  retain_metadata_cols = c("dose_group", "dose")
)
mean_mf <- mf_data_global %>%
  dplyr::group_by(dose_group) %>%
  dplyr::summarise(mean_mf_min = mean(mf_min),
                   SE = sd(mf_min) / sqrt(dplyr::n()))
```

**Table 8.** Mean Mutation Frequency per Dose Group.

```{r mf-mean-table, echo=FALSE}
DT::datatable(mean_mf,
  options = list(
    columnDefs = list(
      list(targets = c("mean_mf_min", "SE"),
           render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }"))
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
)
```

## Mutation Subtypes

Mutations can also be grouped by mutation subtype at varying degrees of
resolution using the `subtype_resolution` parameter. 

Table: (\#tab:subtypes) Definitions of mutation subtype resolutions.

|Subtype resolutions| Definition | Example |
|-------------------|------------|---------|
| `type` | The variation type | snv, mnv, insertion, deletion, complex, and symbolic variants |
| `base_6` | The simple spectrum; snv subtypes in their pyrimidine context | C>A, C>G, C>T, T>A, T>C, T>G |
| `base_12` | The snv subtypes | A>C, A>G, A>T, C>A, C>G, C>T, G>A, G>C, G>T, T>A, T>C, T>G |
| `base_96` | The trinculeotide spectrum; the snv subtypes in their pyrimidine context alongside their two flanking nucleotides | A[C>T]A |
| `base_192` | The snv subtypes reported alongside their two flanking nucleotides | A[G>A]A |

Mutations and `total_depth` will be summed across groups for each mutation
subtype to calculate frequencies. For SNV subtypes, the `total_depth` is
summed based on the sequence context in which the SNV subtype occurs
(`subtype_depth`). In the simplest example, for the `base_6` SNV subtypes, the
two possible reference bases are C or T; hence, the `total_depth` is summed
separately for C:G positions and T:A positions. Thus, the MF for C>T mutations
is calculated as the total number of C>T mutations divided by total_depth for
C:G positions within the group: `sum` / `subtype_depth`. Non-snv mutation
subtypes, such as mnvs, insertions, deletions, complex variants, and structural
variants, will be calculated as their `sum` / `group_depth`, since they can
occur in the context of any nucleotide.  
Upon import of mutation data, columns are created that facilitate the grouping
of SNV subtypes and their associated sequence context for the various
resolutions. Below the columns associated with each subtype_resolution are
defined:

Table: (\#tab:subtypes-cols) Subtype Resolutions and their associated subtype/context columns.

|Subtype resolution | Subtype Column | Context Column |
|-------------------|----------------|----------------|
| `type` | `variation_type`| NA |
| `base_6` | `normalized_subtype` | `normalized_ref` |
| `base_12` | `subtype` | `short_ref` |
| `base_96` | `normalized_context_with_mutation` | `normalized_context` |
| `base_192` | `context_with_mutation` | `context` |

The function will also calculate the proportion of mutations for each subtype,
normalized to the `total_depth`:  
$$P_s = \frac{\left(\frac{M_s}{D_s}\right)}{\sum_s \left(\frac{M_s}{D_s}\right)}$$  
Where $P_s$ is the normalized mutation proportion for subtype $s$. $M_s$ is
the group mutation sum for subtype $s$. $D_s$ is the group sum of the
`subtype_depth` for subtype $s$.  
If total_depth is not available for the mutation data, `calculate_mf()`
will return the subtype mutation counts per group. It will also calculate
subtype proportions, without normalizing to the total_depth:  
$$P'_s = \frac{M_s}{M_{total}}$$  
Where, $P'_s$ is the non-normalized mutation proportion of subtype $s$. $M_s$
is the group mutation sum for subtype $s$. $M_{total}$ is the total mutation
sum for the group.

*Example 3.4. The following code will return the base_6 mutation spectra for*
*all samples with mutation proportions normalized to depth.*

```{r mf-6}
mf_data_6 <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "sample",
  subtype_resolution = "base_6"
)
```

**Table 9.** Frequency and Proportions of 6-base Mutation Subtypes per Sample.

```{r mf-6-table, echo=FALSE}
DT::datatable(mf_data_6,
  options = list(
    scrollX = TRUE,
    columnDefs = list(
      list(targets = c("mf_min", "mf_max"),
           render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }"))
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
)
```

### Selecting Variation Types

`calculate_mf()` can be used on a user-defined subset of `variation_type`
values. The `variant_types` parameter can be set to a character string of
`variation_type` values that the user wants to include in the mutation counts.
When calculating group mutation sums, only variants of the specified
`variation_type`s will be counted. The `total_depth` for records of excluded
`variation_types` will still be included in the `group_depth` and the
`subtype_depth`, if applicable.

By default the function will calculate summary values based on all mutation
types.

*Example 3.5. Calculate global mutation frequencies per sample including only*
*insertion and deletion mutations in the count.*

```{r mf-indels}
mf_data_global_indels <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "sample",
  subtype_resolution = "none",
  variant_types = c("insertion", "deletion")
)
```

**Table 10.** Indel Mutation Frequency per Sample.

```{r mf-indels-table, echo=FALSE}
DT::datatable(mf_data_global_indels,
  options = list(
    columnDefs = list(
      list(targets = c("mf_min", "mf_max"),
           render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }"))
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
)
```

Users may also supply a list of variation_types to exclude to the
`variant_types` parameter, so long as the value is preceeded by a "-".

*Example 3.6. Calculate mutation frequencies at the "type" subtype resolution,*
*excluding ambiguous and uncategorized mutations.*

```{r mf-types}
mf_data_types <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "sample",
  subtype_resolution = "type",
  variant_types = c("-ambiguous", "-uncategorized")
)
```

**Table 11.** Frequency and Proportions of Variation Types per Sample.

```{r mf-types-table, echo=FALSE}
DT::datatable(mf_data_types,
  options = list(
    scrollX = TRUE,
    columnDefs = list(
      list(targets = c("mf_min", "mf_max"),
           render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }"))
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
)
```

*Example 3.7. Include only snv mutations at the base_96 resolution*

```{r mf-96}
mf_data_96 <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "sample",
  subtype_resolution = "base_96",
  variant_types = "snv"
)
```

**Table 12.** Frequency and Proportions of 96-base Trinucleotide Subtypes per Sample.

```{r mf-96_tables, echo=FALSE}
DT::datatable(mf_data_types,
  options = list(
    scrollX = TRUE,
    columnDefs = list(
      list(targets = c("mf_min", "mf_max"),
           render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }"))
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
)
```

## Correcting Depth

There may be cases when the same genomic position is represented multiple times
within the mutation data of a single sample. For instance, we require that
multiple different alternate alleles for a single position be reported as
seperate rows. Another common example of this that we have observed are deletion
calls. Often, for each deletion called, a no_variant call at the same start
position will also be present. For data that includes the `total_depth` of each
record, we want to prevent double-counting the depth at these positions during
when summing the depth across groups. When set to TRUE, `correct_depth` parameter
will correct the depth at these positions. For positions with the same sample,
contig, and start values, the `total_depth` will be retained for only one row.
All other rows in the group will have their `total_depth` set to 0 before beign
summed. The import functions will automatically check for duplicated rows and
return a warning advising to correct the depth should it find any in the mutation
data. This is recommended for all users whose data contain `total_depth` values.
By default, correct_depth is set to TRUE.

*NOTE: The case of deletions and no_variants: When identifying a deletion,*
*some variant callers will re-align the sequences. As such a second*
*total_depth value will be calculated, specifically for the deletion. The*
*variant caller will then provide both the no_variant and the deletion call,*
*the former with the initial depth, and the latter with the re-aligned depth*
*value. Generally, when correcting the depth, the function will retain the*
*total_depth of the first row in the group, and set the rest to 0. However, in*
*the case of deletions, the function may prioritize retaining the re-aligned*
*total_depth over the total_depth assigned to the no_variant. This*
*prioritization can be activated with the `correct_depth_by_indel_priority`*
*parameter.*

## Precalculated Depth

For mutation data that does not include a `total_depth` value for each
sequenced site, precalculated depths for the specified groups can be supplied
separately to `calculate_mf()` through the `precalc_depth_data` parameter.
Depth should be provided at the correct subtype resolution for each level of
the specified grouping variable. This parameter will accept either a data frame
or a filepath to import the depth data. Required columns are:

1. The user-defined cols_to_group variable(s)
2. `group_depth`: the `total_depth` summed across the cols_to_group.
3. The context column for the specified `subtype_resolution`. Only applicable
if using the SNV resolutions (base_6, base_12, base_96, base_192). Column names
are listed in the table above.
4. `subtype_depth`: the `total_depth` summed across the cols_to_group for each
context. Only applicable if using the SNV resolutions.

*Example 3.8. Use precalculated depth values to calculate the global per sample MF.*

**Table 13. ** 'sample_depth' - Precalculated Depth Data per sample.

```{r precalc1-depth}
sample_depth <- data.frame(
  sample = unique(example_data$sample),
  group_depth = c(565395266, 755574283, 639909215, 675090988, 598104021,
                  611295330, 648531765, 713240735, 669734626, 684951248,
                  716913381, 692323218, 297661400, 172863681, 672259724,
                  740901132, 558051386, 733727643, 703349287, 884821671,
                  743311822, 799605045, 677693752, 701163532)
)

DT::datatable(sample_depth)
```

```{r mf-precalc-global}
mf_data_global_precalc <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "sample",
  subtype_resolution = "none",
  calculate_depth = FALSE,
  precalc_depth_data = sample_depth
)
```

**Table 14.** Mutation Frequency per Sample, Precalculated Depth Example.

```{r mf-precalc1-table, echo=FALSE}
DT::datatable(mf_data_global_precalc,
  options = list(
    columnDefs = list(
      list(targets = c("mf_min", "mf_max"),
           render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }"))
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
)
```

*Example 3.9. Use precalculated depth values to calculate the base_6 per sample MF.*

```{r mf-precalc-6}
mf_data_6_precalc <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "sample",
  subtype_resolution = "base_6",
  calculate_depth = FALSE,
  precalc_depth = eh[["EH9862"]]
)
```

**Table 16.** Precalculated Depth Data per sample for base_6 subtype resolution.

```{r precalc2-depth, echo=FALSE}
depth <- eh[["EH9862"]]

DT::datatable(depth)
```

**Table 17.** Frequency and Proportions of 6-base Subtypes per Sample, Precalculated Depth Example.

```{r mf-precalc2-table, echo=FALSE}
DT::datatable(mf_data_6_precalc,
  options = list(
    scrollX = TRUE,
    columnDefs = list(
      list(targets = c("mf_min", "mf_max"),
           render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }"))
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
)
```

View examples for base_12, base_96, and base_192 below:
```{r pre-calc-other-ex}
base_12 <- eh[["EH9863"]]
base_96 <- eh[["EH9864"]]
base_192 <- eh[["EH9865"]]
```

## Summary Table

The function will output the resulting `mf_data` as a data frame with the MF
and proportion calculated. If the `summary` parameter is set to `TRUE`, the
data frame will be a summary table with the MF calculated for each group. If
`summary` is set to `FALSE`, the MF will be appended to each row of the
original `mutation_data`.

The summary table will include:

- `cols_to_group`: all columns used to group the data.
- Subtype column for the given resolution, if applicable.
- Context column for the given resolution, if applicable.
- `sum_min` & `sum_max`: the min/max mutation counts for the group(s)/subtypes.
- `group_depth`: the total_depth summed across the group(s).
- `subtype_depth`: the total_depth summed across the group(s), for the given
context, if applicable.
- `MF_min` & `MF_max`: the min/max MF for the group(s)/subtype.
- `proportion_min` & `proportion_max`: the min/max subtype proportion, if
applicable.

Additional columns from the orginal mutation data can be retained using the
`retain_metadata_cols` parameter. Retaining higher-order experimental groups
may be useful for later statistical analyses or plotting. See above for example
*calculating mean MF per dose*.

**NOTE** The summary table uses a pre-defined list of possible subtypes for each
resolution. If a particular subtype within a given group is not recorded in the
mutation data, the summary table will have no frame of reference for populating
the `metadata_cols`. Thus, for subtypes that do not occur in the mutation data for
a given group, the corresponding `metadata_col` will be NA.

## MF Plots {.tabset}

You can visualize the results of `calculate_mf` using the `plot_mf()` and
`plot_mean_mf`() functions. These functions offer variable aesthetics for
easy visualization. The output is a ggplot object that can be modified using
ggplot2.

### plot_mf

*Example 3.10. Plot the Min and Max MF per sample, coloured and ordered by*
*dose group. See example 3.1 for calculating mf_data_global*

```{r plot-mf-caption, include=FALSE}
caption <- paste(
  "Mutation Frequency (MF) Minimum and Maximum (mutations/bp) per Sample.",
  "Light colored bars represent MFmin and dark coloured bars represent MFmax.",
  "Bars are coloured and grouped by dose.",
  "Data labels are the number of mutations per sample."
)
```

```{r plot-mf, fig.width=12, fig.height=6, fig.cap=caption}
# Define the order for dose groups
mf_data_global$dose_group <- factor(
  mf_data_global$dose_group,
  levels = c("Control", "Low", "Medium", "High")
)
plot <- plot_mf(
  mf_data = mf_data_global,
  group_col = "sample",
  plot_type = "bar",
  mf_type = "both",
  fill_col = "dose_group",
  group_order = "arranged",
  group_order_input = "dose_group"
)
plot
```

### plot_mean_mf()

Calculate and plot the mean MF for a user-defined group using
`plot_mean_mf()`.

*Example 3.11. Plot the mean MF min per dose, including SEM*
*and individual values coloured by dose. See example 3.1 for calculating*
*mf_data_global*

```{r plot-mean-mf-caption, include=FALSE}
caption <- paste(
  "Mean Mutation Frequency (MF) Minimum per Dose. Lines are mean ± S.E.M.",
  "Points are individual samples, coloured by dose."
)
```

```{r plot-mean-mf, fig.cap=caption}
plot_mean <- plot_mean_mf(
  mf_data = mf_data_global,
  group_col = "dose_group",
  mf_type = "min",
  fill_col = "dose_group",
  add_labels = "none",
  group_order = "arranged",
  group_order_input = "dose_group",
  plot_legend = FALSE,
  x_lab = "Dose Group"
)
plot_mean
```

# Exporting Results

Users can easily output data frames to an Excel workbook with `write_excel()`.
This function can write single data frames or it can take a list of dataframes
and write each one to a separate Excel sheet in a workbook.

In addition to data frames, `write_excel()` will also extract the mf_data,
point_estimates, and pairwise_comparisons from `model_mf()` output to write to
an excel workbook. Set `model_results` to TRUE if supplying the function with
the output to model_mf(). See Modelling Mutation Frequencies for more information
on using `model_mf()`.

*Example 4.1. Write MF data to excel workbook.*
```{r write-excel, eval=FALSE}
# save a single data frame to an Excel file
write_excel(mf_data_global, workbook_name = "example_mf_data")

# Write multiple data frames to a list to export all at once.
list <- list(mf_data_global, mf_data_rg, mf_data_6)
names(list) <- c("mf_per_sample", "mf_per_region", "mf_6spectra")

#save a list of data frames to an Excel file
write_excel(list, workbook_name = "example_mf_data_list")

```


Mutation data can be written to a VCF file for downstream applications with
`write_vcf_from_mut()`.

*Examples 4.2: export our example data as a vcf file.*

```{r write-vcf, eval=FALSE}
write_vcf_from_mut(example_data)
```

# Summary Report

To facilitate reproducible analysis of error-corrected sequencing data, the
MutSeqR package includes a pre-built R Markdown workflow: Summary_Report.Rmd.
This report template allows users to efficiently evaluate the effects of an
experimental variable (e.g., dose, age, tissue) on mutation
frequency and mutational spectra.

The workflow guides users through a standardized analysis pipeline that
includes:

- Mutation data Import
- Mutation data Filtering
- Calculation of Global Mutation Frequencies per Sample
- Visualization of mutation frequencies per Sample and Mean mutation
frequencies per experimental group.
- Generalized Linear Modeling of Mutation Frequency (min) as an effect of
the experimental variable.
- Optional Benchmark Dose Modeling of mutation frequency across a dose or
concentration variable.
- Calculation of Mutation Subtype Frequencies and Proportions at the base-6
and base-96 resolutions.
- Comparison of base-6 mutation spectra between experimental groups.
- Visualization of base-6 subtype proportions per sample with hierarchical
clustering
- Visualization of base-6 subtype proportions per experimental group.
- Visualization of base-96 subtype proportions per experimental group.
- Mutation signature analysis per experimental group.
- Generalized Linear Mixed Modeling of mutation frequency as an effect of
specified genomic regions and the experimental group.
- Visualization of multiplet mutations per experimental group.

All figures will be included in the report. Data frames included in the report
will also be exported to excel workbooks within a predefined output path.

## General Usage

Users must download the config file "summary_config.yaml". This file defines
the parameters to be passed to the Summary_Report.rmd. Users must fill out
the .yaml file and save it.

```{r download-yaml, eval=FALSE}
config <- system.file("extdata", "inputs", "summary_config.yaml", package = "MutSeqR")
file.copy(from = config, to = "path/to/save/the/summary_config.yaml")
```

Users can then render the report using MutSeqR's render_report().
Provide the function with the .yaml file using the `config_file_path`
parameter. Provide the name and the file path of the output file to the
`output_file` parameter. Set the `output_format` to one of "html_document"
(recommended), "pdf_document", or "all".

```{r render-report, eval=FALSE}
render_report(config_file_path = "path/to/summary_config.yaml",
              output_file = "path/to/output_file.html",
              output_format = "html_document")
```

## Parameters

Below is a description of the parameters set in summary_config.yaml.
Parameters are catagerised by their function.

System Set Up

- *projectdir*: The working directory for the project. All filepaths will be
imported from the projectdir. If NULL, the current working directory will be
used.
- *outputdir*: The output directory where output files will be saved. If NULL,
the projectdir will be used.

Project Details

- *project_title*: The project title.
- *research_name*: The name of the principle investigator for the project.
- *user_name*: The name of the one running the analysis.
- *project_description*: Optional description of the project.

Workflow Profile

- *config_profile* : Choose a workflow profile to load pre-defined parameters.
Current options are "Duplex Sequencing Mouse Mutagenesis Panel",
"Duplex Sequencing Human Mutagenesis Panel", 
"Duplex Sequencing Rat Mutagenesis Panel", "CODEC"^1, and "None". Pre-defined
parameters are set to handle data import and filtering. More profiles coming
soon. If not using a pre-defined profile ("None"), users must fill out the
Custom_Profile_Params (see below).

^1 CODEC .maf files output from the CODECsuite analysis pipeline may be
imported provided the user 1. Adds a "sample" column, 2. Adds an "end" column
equal to the Start_Position + length(Reference_Allele) - 1, 3. Ensure Chromsome
entries are consistent with notation of BSgenome seqnames for the given reference
genome. Ex. GRCh38 requires users to remove "chr" from the Chromosome value.

Data Import

- *BS_genome*: The BS genome to use to populate the sequence context. The given
BS genome must be installed prior to running the render_report() function.
To identify an appropriate BS genome based on species and genome assembly,
use find_BS_genome(). Ex. "BSgenome.Mmusculus.UCSC.mm10"
- *file_type*: The format of the mutation data to be imported. Options are table
or vcf.
- *mutation_file*: The file path to the mutation data file(s). This can be an
individual file or a directory containing multiple files. The file path will be
read as projectdir/mutation_file.
- *mut_sep*: The delimiter for importing tabular mutation data. Ex. "\t" is
tab-delimited.
- *sample_data*: An optional file path to the sample metadata that will be joined
with the mutation data. Set to NULL if not importing metadata. The file path will
be read as projectdir/sample_data.
- *sd_sep*: The delimiter for importing the sample_data file. Ex. "\t" is
tab-delimited.

Calculating MF

Precalculated-depth files: The Summary_Report will automatically check during
import whether the mutation data contains a depth column (total_depth or
depth). If a depth column does not exist, per-sample precalculated depth can be
provided for any subtype resolution to calculate mutation frequencies. If an
exp_variable is provided, the Summary_Report will automatically sum the
per-sample depths to obtain the depth per experimental group as needed.

*In order for MutSeqR to calculate the depth from the mutation data, the*
*data must have the depth-value for **every sequenced site**. It is not*
*sufficient to calculate the depth from a list of variants. No-variant sites*
*must be included. The Summary Report will check for a total_depth column, but it will*
*not check whether no_variant sites are included. If it finds a total_depth*
*column, it will proceed to calculate the depth from the mutation data*.


- *precalc_depth_data_global*: Optional file path to the precalculated per-sample
total_depth data. This is the total number of bases sequenced per sample, used
for calculating mutation frequencies. Columns are "sample" and "group_depth".
If using an exp_variable (see below), please also include it in this table.
The file path will be read as projectdir/precalc_depth_data_global.
- *precalc_depth_data_base6*: Optional file path to the precalculated per-sample
total_depth data in the base_6 context. This is the total number of C and T
bases sequenced for each sample. Columns are "sample", "normalized_ref", and
"subtype_depth". If using an exp_variable (see below), please also include it
in this table. The file path will be read as
projectdir/precalc_depth_data_base6.
- *precalc_depth_data_base96*: Optional file path to the precalculated
per-sample total_depth data in the base_96 context. This is the total
number bases sequenced per sample for each of the 32 possible trinucleotide
contexts in their pyrimidine notation. Columns are "sample",
"normalized_context", and "subtype_depth". If using an exp_variable,
please also include it in this table. The file path will be read as
projectdir/precalc_depth_data_base96. 
- *precalc_depth_data_rg*: Optional file path to the precalculated
per-sample total_depth data for each target region. This is the total
number of bases sequenced per sample for each region. Columns are "sample",
'region_col', "group_depth". Only applicable if performing  regions analysis.
The file path will be read as projectdir/precalc_depth_data_rg.
- *d_sep*: The delimiter for importing the precalc_depth_data files.

Statistical Analysis

The Summary Report will run several analyses to investigate the effect
of a specified experimental variable on mutation frequency and spectra.

- *exp_variable*: Optional. The experimental variable of interest. This
argument should refer to a column in your mutation data or sample_data.
The Summary Report does not support the analysis of multiple exp_variables;
supply only one column name. Ex. "dose". If NULL, statistical analyses will
be skipped.
- *exp_variable_order*: A vector specifying the unique levels of the
exp_variable in the order desired for plotting.
- *reference_level*: The reference level of the exp_variable. Ex. the vehicle
control group for a chemical dose. This value must match one of the levels
within the exp_variable column. If your experimental variable does not have a
reference level, then any level of the variable is fine.
- *contrasts*: An optional file path to the contrasts table specifying
pairwise comparisons between levels of the experimental variable.
Required 2 columns, no header. The level in the first column will be compared
to the level in the second column for each row. P-values will be adjusted for 
multiple comparisons if more than one comparison is specified. This contrasts
table will be used for Generalized Linear Modeling of MFmin by the exp_variable
(model_mf), Generalized Linear Mixed Modeling of MFmin by the exp_variable and
target regions, if using a targeted panel (model_mf), and comparing the base_6
mutation spectra between exp_variable levels (spectra_comparison). The file
path will be read as projectdir/contrasts. If NULL, spectra_comparison will be
skipped, and the GLM/GLMM will return only the model estimates per exp_variable.
- *cont_sep*: the delimiter for importing the contrasts file.
- *bmd*: A logical variable indicating whether to run a BMD analysis on
the MF data. If TRUE, the exp_variable must refer to numeric dose or
concentration values.
- *bmr*: A numeric value indicating the benchmark response for the bmd analysis.
The Summary Report defines the bmr as a bmr-% increase in MF from the reference
level. Default is 0.5.

Custom Profile Parameters

Users who do not wish to use one of the pre-defined parameter profiles may
still run the Summary Report by filling out the Custom_Profile_Params.
If using one of the config_profiles, users should skip this section.

- *ecs_technology*: The technology used to generate the data.
- *is_0_based_mut*: A logical variable indicating whether the genomic
coordinates of the tabular mutation data are 0-based (TRUE), or 1-based (FALSE).
- *regions*: An optional file path to the regions metadata file. The file must
contain columns "contig", "start", "end" in addition to the metadata.
- *is_0_based_rg*: A logical variable indicating whether the genomic
coordinates of the regions file are 0-based (TRUE), or 1-based (FALSE).
- *rg_sep*: The delimiter for importing the regions file.
- *vaf_cutoff*: A numeric value from 0-1. Will flag variants with a VAF >
vaf_cutoff as germline variants and filter them from mutation counts.
- *snv_in_germ_mnv*: A logical variable indicating whether filter_mut should
flag SNVs that overlap with germline MNVs for filtering.
- *rm_abnormal_vaf*: A logical variable indicaring whether filter_mut should
remove records where the VAF is between 0.05-0.45 or between 0.55-0.95.
- *custom_filter_col*: A column name used by filter_mut to apply a custom filter
to the specified column.
- *custom_filter_val*: The value in custom_filter_col by which to filter.
- *custom_filter_rm*: A logical variable indicating whether records that
contain the custom_filter_val within the custom_filter_col should be removed
(TRUE) or flagged (FALSE).
- *filtering_regions*:  Optional file path to regions file by which to filter
variants. Must contain the contig, start, and end of each region. The file
path will be read as projectdir/filtering_regions.
- *filtering_rg_sep*: The delimiter for importing the filtering_regions file.
- *regions_filter*: "keep_within" will remove any records that fall outside of
the filtering_regions. "remove_within" will remove records that fall inside of
the filtering_regions.
- *allow_half_overlap*: A logical variable indicating whether to include records
that half overlap with the regions. If FALSE, the start and end position of the
record must fall within the region interval to be counted as "falling in the
region". If TRUE, records that start/end within the region interval, but extend
outside of it will be counted as "falling inside the region".
- *filtering_is_0_based_rg*: A logical variable indicating whether the genomic
coordinates of the filtering_regions are 0-based (TRUE) or 1-based (FALSE).
- *rm_filtered_mut_from_depth*: A logical variable indicating whether the
alt_depth of variants flagged during the filtering process should be removed
from their total_depth values. This does not apply to records flagged as
germline variants.
- *do_regions_analysis*: A logical variable indicating whether to perform
Generalized Linear Mixed Modeling of the data by sequencing target. This is
applicable to data sets that used a targeted panel or has specific regions
of interest.
- *region_col*: The column name that uniquely identifies each target region
for the regions_analysis. This column must be present in the mutation_data
or in the regions metadata file.

## Output

render_report() will output the summary report of results as an html file or a pdf file.
All output will be saved to the output_directory specified in the summary_config.yaml.

### Report

The Summary Report consists of the following sections.

1. **Import the Mutation Data:** Statement of number of imported rows and the
total number of samples detected.
2. **Filter the Mutation Data**: Summary of applied filters with total counts
for the number of records that were flagged/removed.
3. **Calculate Per-Sample Mutation Frequency**: Mutation frequency calculated
per sample. Includes plot of mutation frequency per sample and plot of mean
mutation frequency per experimental group (given experimental variable was
provided).
4. **Generalized Linear Modelling**: Model results of mutation frequency as an
effect of the experimental variable. Includes model summary, residual plots,
model estimates, pairwise comparisons as specified in "contrasts", and a plot
of the model estimates and significant results.
5. **Benchmark Dose Analysis**: BMD results for the experimental variable on
mutation frequency. BMD estimates per individual model and model-averaged
result. Includes model plots, cleveland plots, and confidence interval plots.
6. **6-Base Spectrum**: Mutation frequency calculated at the base_6 resolution
per-sample and averaged across the experimental variable. Includes hierarchical
clustering of samples by subtype proportion, spectra comparison between experimental
groups, and plot of subtype proportions per experimental group.
7. **96-Base Trinucleotide Spectrum**: Mutation frequency calculated at the base_96
resolution per-sample and averaged across the experimental variable. Includes
trinucleotide proportion plots for all experimental groups or by sample if no
experimental variable is provided.
8. **Regions Analysis**: Mutation frequency calculated per sample for genomic
regions of interest. Includes model results of mutation frequency as an effect
of the experimental variable and genomic target. Pairwise comparisons specified
in contrasts are conducted for each genomic target individually. If no
experimental variable is provided, model results reflect mutation frequency as
an effect of genomic target. 
9. **Mutation HotSpots**: Statement # of mutations that occur in more than one sample.
Also includes recurrent mutations within experimental groups.
10. **Lollipop Plots**: Given genomic regions of interest, plots recurrent mutations
across genomic position.
11. **Bubble Plot: Multiple Mutations**: Bubble plot depicting multiplet mutations per
experimental group.

If an experimental variable is not provided, statistical analysis will be skipped.


# References

# Appendix

## Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```
