---
title: "MutSeqR: Analyzing Mutation Spectra"
author:
  - name: Annette E. Dodge
    affiliation: Environmental Health Science and Research Bureau, Health Canada, Ottawa, ON, Canada.
  - name: Matthew J. Meier
    affilitation: Environmental Health Science and Research Bureau, Health Canada, Ottawa, ON, Canada.
    email: matthew.meier@hc-sc.gc.ca
package: MutSeqR
output:
  rmarkdown::html_document:
    toc: true
    toc_float: yes
    code_folding: show
    code_download: yes
    css: custom.css
bibliography: Analyzing_Mutation_Spectra.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(DT)
library(htmltools)
```

The mutation spectra is the pattern of mutation subtypes within a sample or
group. The mutation spectra can inform on the mechanisms involved in
mutagenesis which is used to extrapolate findings across diverse populations
and species.

# Load MutSeqR and Example Data

```{r load-lib}
library(MutSeqR)
```

```{r load-ex}
library(ExperimentHub)
# load the index
eh <- ExperimentHub()
```

# Comparing Mutation Spectra between Groups

The mutation spectra can be compared between experimental groups using the
`spectra_comparison()` function. This function will compare the proportion
of mutation subtypes at any resolution between user-defined groups using a
modified contingency table approach [@piegorsch-1994].

This approach is applied to the mutation counts of each subtype in a given group. The contingency table is represented as $R * T$ where $R$ is the number of subtypes, and $T$ is the number of groups. `spectra_comparison()` performs comparisons between $T = 2$ specified groups. The statistical hypothesis of homogeneity is that the proportion (count/group total) of each mutation subtype equals that of the other group. To test the significance of the homogeneity hypothesis, the $G^{2}$ likelihood ratio statistic is used:

$$G^{2} = 2\  \sum_{i=1}^{R}\  \sum_{j=1}^{T}\  Y_{ij}\  log(\frac{Y_{ij}}{E_{ij}})$$

$Y_{ij}$ represents the mutation counts and $E_{ij}$ are the _expected_ counts
under the null hypothesis. The $G^{2}$ statistic possesses approximately a
$\chi^{2}$ distribution in large sample sizes under the null hypothesis of
no spectral differences. Thus, as the column totals become large, $G^{2}$ may
be referred to a $\chi^{2}$ distribution with $(R -  1)(T - 1)$ degrees of
freedom. The $G^{2}$ statistic may exhibit high false positive rates in small
sample sizes when referred to a $\chi^{2}$ distribution. In such cases, we
instead switch to an F-distribution. This has the effect of reducing the rate
at which $G^{2}$ rejects each null hypothesis, providing greater stability in
terms of false positive error rates. Thus when $N/(R-1) < 20$, where $N$ is
the total mutation counts across both groups, the function will use a
F-distribution, otherwise it will use a $\chi^{2}$-distribution.

This comparison assumes independance among the observations. Each tabled
observation represents a sum of independent contributions to the total mutant
count. We assume independance is valid for mutants derived from a mixed
population, however, mutants that are derived clonally from a single progenitor
cell would violate this assumption. As such, it is recommended to use the
**MFmin method** of mutation counting for spectral analyses to ensure that
all mutation counts are independant. In those cases where the independence
may be invalid, and where additional, extra-multinomial sources of variability
are present, more complex, hierarchical statistical models are required. This
is outside the scope of this package.

### General Usage: spectra_comparison()

The first step is to calculate the per-group mf data at the desired
subtype_resolution using `calculate_mf()`. This mf_data is then supplied
to `spectra_comparison()` along with a contrasts table to specify the
comparisons.

Specify the variable(s) by which you wish to make your comparisons with The
`exp_variable` parameter. Variable(s) must be present in your mf_data.

The `contrasts` parameter can be supplied with either a data
frame or a filepath to a text file that will be loaded into R. The table must
consist of two columns, each containing levels within the fixed_effects.
The level in the first column will be compared to the level in the second
column for each row. If you are comparing across more than one experiment
variable, seperate the levels of the variables you are comparing with a
colon. A level from each variable must be included in all values of the
contrasts table.

The function will output the $G^{2}$ statistic and p-value for each specified
comparison listed in `constrasts`. P-values are adjusted for multiple comparison using the Sidak method (**adj_p.value**).

_Example 1. In our example data, we are studying the mutagenic effect of_
_BaP. Our samples were exposed to three doses of a BaP (Low, Medium, High),_
_or to the vehicle control (Control). We will compare the base_6 snv subtypes,_
_along with non-snv variants, of each of the three dose groups to the control._
_In this way we can investigate if exposure to BaP leads to significant spectral_
_differences._

First, we will calculate the per-dose MF at the 6-based subtype resolution.
This function will use the mutation sums (sum_min) that are calculated for
each subtype per dose group, not the frequencies.

```{r calculate-mf-6}
# load example data:
example_data <- eh[["EH9861"]]

# Calculate the MF per dose at the base_6 resolution.
mf_data_6_dose <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "dose_group",
  subtype_resolution = "base_6"
)
```

Next, we generate a contrasts table that specifies the pairwise comparisons we
want to perform. We will compare each BaP dose group to the control group.

```{r contrasts1}
# Create the contrast table
contrasts_table <- data.frame(
  col1 = c("Low", "Medium", "High"),
  col2 = c("Control", "Control", "Control")
)
```

Finally, run the analysis.

```{r spectra-comp6}
# Run the analysis
ex_spectra_comp <- spectra_comparison(
  mf_data = mf_data_6_dose,
  exp_variable = "dose_group",
  contrasts = contrasts_table
)
```

**Table 23.** Comparison of the 6-base subtype Proportions between Dose Groups.

```{r spectra-comp-table, echo=FALSE}
DT::datatable(
  ex_spectra_comp,
  options = list(
    columnDefs = list(
      list(targets = c("p.value", "adj_p.value"),
           render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }"))
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
) %>%
  DT::formatRound(columns = "G2", digits = 2)
```

Our results indicate the base-6 mutation spectra of all three BaP dose groups are significantly different from that of the control group.

## Mutational Signatures

Mutational processes generate characteristic patterns of mutations, known as
mutational signatures. Distinct mutational signatures have been extracted from
various cancer types and normal tissues using data from the Catalogue of Somatic
Mutations in Cancer, ([COSMIC](https://cancer.sanger.ac.uk/signatures/))
database. These include signatures of single base substitutions (SBSs),
doublet base substitutions (DBSs), small insertions and deletions (IDs) and
copy number alterations (CNs). It is possible to assign mutational signatures
to individual samples or groups using the `signature_fitting` function. Linking
ECS mutational profiles of specific mutagens to existing mutational signatures
provides empirical evidence for the contribution of environmental mutagens to
the mutations found in human cancers and informs on mutagenic mechanisms.

The `signature_fitting` function utilizes the
[SigProfiler](https://github.com/AlexandrovLab) suite of tools
(@diaz-gay-2023; @khandekar-2023) to assign SBS signatures from
the COSMIC database to the 96-base SNV subtypes of a given group by creating
a virtual environment to run python using `reticulate`. `signature_fitting()`
facilitates interoperability between these tools for users less familiar with
python and assists users by coercing the mutation data to the necessary
structure for the SigProfiler tools.

This function will install several python dependencies using a conda virtual
environment on first use, as well as the FASTA files for all chromosomes for
your specified reference genome. As a result ~3Gb of storage must be available
for the downloads of each genome.

SNVs in their 96-base trinucleotide context are summed across groups to create
a mutation count matrix by `SigProfilerMatrixGenerator()`
(SigProfilerMatrixGenerator; @khandekar-2023). `Analyze.cosmic_fit `
(SigProfilerAssignment; @diaz-gay-2023) is then run to assign mutational
signatures to each group using refitting methods, which quantifies the
contribution of a set of signatures to the mutational profile of the group.
The process is a numerical optimization approach that finds the combination of
mutational signatures that most closely reconstructs the mutation count matrix.
To quantify the number of mutations imprinted by each signature, the tool uses
a custom implementation of the forward stagewise algorithm and it applies
nonnegative least squares, based on the Lawson-Hanson method. Cosine similarity
values, and other solution statistics, are generated to compare the
reconstructed mutational profile to the original mutational profile of the
group, with cosine values > 0.9 indicating a good reconstruction.

Currently, `signature_fitting` offers fitting of COSMIC version 3.4 SBS
signatures to the SBS96 matrix of any sample/group. For advanced use, including
using a custom set of reference signatures, or fitting the DBS, ID, or CN
signatures, it is suggested to use the SigProfiler python tools directly in as
described in their respective [documentation](https://github.com/AlexandrovLab).

### General Usage: signature_fitting()

_Examples include external dependencies_

`signature_fitting` requires the installation of reticulate
as well as a version of python 3.8 or newer.

```{r install-sigprof-depends, eval=FALSE}
# Install reticulate
install.packages("reticulate")

# Install python
reticulate::install_python()

```

`signature_fitting` will take the imported (and filtered, if applicable)
`mutation_data` and create mutational matrices for the SNV mutations.
Mutations are summed across levels of the `group` parameter. This can be set to
individual samples or to an experimental group. Variants flagged in the
`filter_mut` column will not be included in the mutational matrices.

The `project_genome` will be referenced for the creation of the mutational
matrices. The reference genome is installed automatically from ENSEMBL via
FTP if it is not already.

The virtual environment can be specified with the `env_name` parameter. If no
such environmnent exists, then the function will create one in which to store
the dependencies and run the signature refitting. Specify your version of python
using the `python_version` parameter (must be 3.8 or higher).

_Example 2. Determine the COSMIC SBS signatures associated with each BaP_
_dose group._

```{r sig-fit, eval=FALSE}
# Run Analysis
signature_fitting(
  mutation_data = example_data, # filtered mutation data
  project_name = "Example",
  project_genome = "mm10",
  env_name = "MutSeqR",
  group = "dose_group",
  python_version = "3.11",
  output_path = NULL
)
```

### Output

Results from the `signature_fitting` will be stored in an output folder.
A filepath to a specific output directory can be designated using the
`output_path` parameter. If null, the output will be stored within your
working directory. Results will be organized into subfolders based on the
`group` parameter. The output structure is divided into three folders: input,
output, and logs.

**Input** folder: "mutations.txt", a text file with the mutation_data
coerced into the required format for SigProfilerMatrixGenerator. It consists
of a list of all the snv variants in each `group` alongside their genomic
positions. This data serves as input for matrix generation.

**Log** folder: the error and the log files for SigProfilerMatrixGeneration.

The output folder contains the results from matrix generation and
signature refitting desribed in detail below.

#### Matrix Generation Output

`signature_fitting` uses
[SigProfilerMatrixGenerator](https://osf.io/s93d5/wiki/home/) to
create mutational matrices ([@bergstrom-2019]). Mutation matrices are
created for single-base substitutions (SBS) and doublet-base substitutions
(DBS), including matrices with extended sequence context and transcriptional
strand bias. SBS and DBS matrices are stored in their respective folders in
the output directory. **Only the SBS96 matrix is used for refitting**.
_Matrices are stored as `.all` files which can be viewed in a text-editor._

Table: (\#tab:mat-files) Output Files from SigProfilerMatrixGeneration by Folder.

| Folder | File                                    | Definition                                                                                                                                                                       | Plot file                          |
| ------ | --------------------------------------- | -------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- | ---------------------------------- |
| SBS    | _project.SBS6.all_                      | The 6 pyrimidine single-nucleotide variants. _C>A, C>G, C>T, T>A, T>C, or T>G_                                                                                                   | _plots/SBS_6_plots_project.pdf_    |
| SBS    | _project.SBS18.all_                     | The 6 pyrimidine single-nucleotide variants within 3 transcriptional bias categories: Untranscribed (U), Transcribed (T), Non-Transcribed Region (N).                            |                                    |
| SBS    | _project.SBS24.all_                     | The 6 pyrimidine single-nucleotide variants within 4 transcriptional bias categories: Untranscribed (U), Transcribed (T), Bidirectional (B), Non-Transcribed Region (N).         | _plots/SBS_24_plots_project.pdf_   |
| SBS    | **project.SBS96.all**                   | The 6 pyrimidine single-nucleotide variants alongside their flanking nucleotides (4 x 4 = 16 combinations). _Ex. A[C>G]T_                                                        | _plots/SBS_96_plots_project.pdf_   |
| SBS    | _project.SBS288.all_                    | The 96-base single-nucleotide variants within 3 transcriptional bias categories (U, T, N).                                                                                       | _plots/SBS_288_plots_project.pdf_  |
| SBS    | _project.SBS384.all_                    | The 96-base single-nucleotide variants within 4 transcriptional bias categories (U, T, N, B).                                                                                    | _plots/SBS_384_plots_project.pdf_  |
| SBS    | _project.SBS1536.all_                   | The 6 pyrimidine single-nucleotide variants alongside their flanking dinucleotides (16 x 16 = 256 combinations). _Ex. AA[C>G]TT_                                                 | _plots/SBS_1536_plots_project.pdf_ |
| SBS    | _project.SBS4608.all_                   | The 1536-base single-nucleotide variants within 3 transcriptional bias categories (U, T, N).                                                                                     |                                    |
| SBS    | _project.SBS6144.all_                   | The 1536-base single-nucleotide variants within 4 transcriptional bias categories (U, T, N, B).                                                                                  |                                    |
| DBS    | _project.DBS78.all_                     | The 78 pyrimidine double-nucleotide variants.                                                                                                                                    | _plots/DBS_78_plots_project.pdf_   |
| DBS    | _project.DBS150.all_                    | The 36 dinucleotide combinations that have only all purines or all pyrimidines x 3 transcriptionla bias categories (U, T, N).                                                    |                                    |
| DBS    | _project.DBS186.all_                    | The 36 dinucleotide combinations that have only all purines or all pyrimidines x 4 transcriptional bias categories (U, T, N, B).                                                 | _plots/DBS_186_plots_project.pdf_  |
| DBS    | _project.DBS1248.all_                   | The 78 pyrimidine double-nucleotide variants alongside their flanking nucleotides (Possible starting nucleotides (4) x 78 x possible ending nucleotides (4) = 1248 combinations) |                                    |
| DBS    | _project.DBS2400.all_                   | The 36 dinucleotide combinations that have only all purines or all pyrimidines alongside their flanking nucleotides within transcriptional bias categories (U, T, N).            |                                    |
| DBS    | _project.DBS2676.all_                   | The 36 dinucleotide combinations that have only all purines or all pyrimidines alongside their flanking nucleotides within 4 transcriptional bias categories (U, T, N, B).       |                                    |
| TSB    | _strandBiasTes_24.txt_                  | Transcription Strand Bias Test stats of the SBS6 variants                                                                                                                        |                                    |
| TSB    | _strandBiasTes_384.txt_                 | Transcription Strand Bias Test stats of the SBS96 variants                                                                                                                       |                                    |
| TSB    | _strandBiasTes_6144.txt_                | Transcription Strand Bias Test stats of the SBS1536 variants                                                                                                                     |                                    |
| TSB    | _significantResults_strandBiasTest.txt_ | Returns significant results from the three files above.                                                                                                                          |                                    |

**Doublet-base Matrices (DBS)**: DBS are somatic mutations in which a set of two
adjacent DNA base-pairs are simultaneously substituted with another set of two
adjacent DNA base-pairs. **We do not recommend using the DBS matrices**
**generated using `signature_fitting`** The `signature_fitting` function is
designed to handle only SBS mutations. All true MNVs, including doublets, are
filtered out of the `mutation_data` prior to MatrixGeneration. However, the
tool will still attempt to identify DBSs and will occasionally find two
independent SBSs occuring next to each other simply by chance. If you wish
to use DBS mutations in your signature analysis, please refer directly to
the SigProfiler tools.

**Plots**: Barplots of the mutation matrices for all groups can be found in
the "plots" folder. The number of mutations are plotted for each group at the
various subtype resolutions. Files for each matrix are listed in
Table \@ref(tab:mat-files).

**vcf_files**: This output folder provides text-based files containing the
original mutations and their SigProfilerMatrixGenerator classification for
each chromosome. The files are separated into dinucleotides (DBS),
multinucleotide substitutions (MNS), and single nucleotide variants (SNV)
folders containing the appropriate files. The headers are:

1. The group
2. the chromosome
3. the position
4. the SigProfilerMatrixGenerator classification
5. the strand {1, 0, -1}.

The headers for each file are the same with the exception of the MNS files
which don't contain a matrix classification or a strand classification.
As noted above the DBS and MNS matrices do no reflect the true mutation
counts for these variant types. Only SBS/SNV mutations are included in the
matrix generation.

**Transcription Strand Bias (TSB)**: SBS and DBS mutations are tested for
[transcription strand bias](https://osf.io/s93d5/wiki/5.%20Output%20-%20TSB/).
Mutations are first classified within the four transcriptional bias
categories:

Table: (\#tab:transcript-cat) SigProfiler Transcriptional Bias Categories.

| Category           | Description                                                   |
| ------------------ | ------------------------------------------------------------- |
| Transcribed (T)    | The variant is on the transcribed (template) strand.          |
| Untranscribed (U)  | The variant is on the untranscribed (coding) strand.          |
| Bidirectional (B)  | The variant is on both strands and is transcribed either way. |
| Nontranscribed (N) | The variant is in a non-coding region and is untranslated.    |

The tool will then perform a transcription strand bias test which compares the
number of transcribed and untranscribed mutations for each mutation type.
_For example, it will compare the number of transcribed T>C to untranscribed_
_T>C mutations. Should there be a significant difference, it would indicate that_
_T:A>C:G mutations are occuring at a higher rate on one of the strands compared_
_to the other._ The output files contain the following headers:

1. the `group`
2. the mutation type
3. the enrichment value (# Transcribed / # untranscribed)
4. the p-value, corrected for multiple comparisons using the false discrover rate method
5. the false discovery rate q-value

#### Signature Refitting Results

Results from the signature refitting perfomed by
[SigProfilerAssignment](https://osf.io/mz79v/wiki/home/) will be stored within
the "Assignment_Solution" folder. "Assignment_Solution" consists of 3
subdirectories; "Activities", "Signatures", and "Solution_Stats".

**Activities**

| File                                     | Description                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                            |
| ---------------------------------------- | ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------ |
| _Assignment_Solution_Activities.txt_     | This file contains the activity matrix for the selected signatures. The first column lists all of the samples/groups. All of the following columns list the calculated activity value for the respective signatures. Signature activities correspond to the specific numbers of mutations from the sample's original mutation matrix caused by a particular mutational process.                                                                                                                                                                        |
| _Assignment_Solution_Activity_Plots.pdf_ | This file contains a stacked barplot showing the number of mutations in each signature on the y-axis and the samples/groups on the x-axis.                                                                                                                                                                                                                                                                                                                                                                                                             |
| _Assignment_Solution_TMB_plot.pdf_       | This file contains a tumor mutational burden plot. The y-axis is the somatic mutations per megabase and the x-axis is the number of samples/groups plotted over the total number of samples/groups included. The column names are the mutational signatures and the plot is ordered by the median somatic mutations per megabase.                                                                                                                                                                                                                      |
| _Decomposed_Mutation_Probabilities.txt_  | This file contains the probabilities of each of the 96 mutation types in each sample/group. The probabilities refer to the probability of each mutation type being caused by a specific signature. The first column lists all the samples/groups, the second column lists all the mutation types, and the following columns list the calculated probability value for the respective signatures.                                                                                                                                                       |
| _SampleReconstruction_                   | This folder contains generated plots for each sample/group summarizing the assignment results. Each plot consists of three panels. (i) Original: a bar plot of the inputted 96SBS mutation matrix for the sample/group. (ii) Reconstructed: a bar plot of the reconstruction of the original mutation matrix. (iii) The mutational profiles for each of the known mutational signatures assigned to that sample/group, including the activities for each signature. Accuracy metrics for the reconstruction are displayed at the bottom of the figure. |

**Signatures**

| Files                                  | Description                                                                                                                                                                                                                                            |
| -------------------------------------- | ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------ |
| _Assignment_Solution_Signatures.txt_   | The distribution of mutation types in the input mutational signatures. The first column lists all 96 of the mutation types. The following columns are the signatures.                                                                                  |
| _SBS_96_plots_Assignment_Solution.pdf_ | Barplots for each signature identified that depicts the proportion of the mutation types for that signature. The top right corner also lists the total number of mutations and the percentage of total mutations assigned to the mutational signature. |

**Solution_Stats**

| Files                                              | Description                                                                                                                                                                                                                                                                                                                                                                        |
| -------------------------------------------------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
| _Assignment_Solution_Samples_Stats.txt_            | The accuracy metrics for the reconstruction. statistics for each sample including the total number of mutations, cosine similarity, L1 norm (calculated as the sum of the absolute values of the vector), L1 norm percentage, L2 norm (calculated as the square root of the sum of the squared vector values), and L2 norm percentage, along with the Kullback-Leibler divergence. |
| _Assignment_Solution_Signature_Assignment_log.txt_ | The events that occur when known signatures are assigned to an input sample. The information includes the L2 error and cosine similarity between the reconstructed and original sample within different composition steps.                                                                                                                                                         |

**Other Files**

_JOB_METADATA_SPA.txt_: This file contains the metadata about system and runtime.

### Use SigProfiler Webtool

Users may choose to use the
[SigProfiler Webtool](https://cancer.sanger.ac.uk/signatures/assignment/)
instead of using the `signature_fitting() `function. MutSeqR offers functions
to coerce mutation data into the proper format for input files.

#### Mutation Calling File

`write_mutation_calling_file()` creates a simple text file from mutation data
that can be used for mutation signatures analysis using the SigProfiler
Assignment web application as a "mutation calling file". Signature analyses
are done at the sample level when using mutation calling files. The file will
be saved to your output directory, specified in `output_path`.

_Example 3. Analyze the COSMIC SBS signatures contributing to each of the_
_24 samples using the SigProfiler Web Tool. Output a mutation calling file that_
_can be uploaded to the webtool._

```{r mut-calling-file, eval=FALSE}
write_mutation_calling_file(
  mutation_data = example_data,
  project_name = "Example",
  project_genome = "mm10",
  output_path = NULL
)
```

#### Mutational Matrix

`write_mutational_matrix()` will sum mutations across user-defined groups
before coercing the data into the proper format for input as a "mutational
matrix". SNV subtypes can be resolved to either the base_6 or base_96
resolution. The file is saved to the specified output directory.

_Example 4. Analyze the COSMIC SBS signatures contributing to each dose_
_group using the SigProfiler Web Tool. Output a mutational matrix that can be_
_uploaded to the webtool._

```{r write-mut-matrix, eval=FALSE}
write_mutational_matrix(
  mutation_data = example_data,
  group = "dose_group",
  subtype_resolution = "base_96",
  mf_type = "min",
  output_path = NULL
)
```

# Visualizing the Mutation Spectra

## plot_spectra {.tabset}

The mutation spectra can be visualized with `plot_spectra` which will create a
stacked bar plot for user-defined groups at the desired subtype resolution.
Mutation subtypes are represented by colour. The value can represent subtype
`proportion`, frequency (`mf`), or count (`sum`).

We will first calculate the mf data at the 6-base resolution for each dose.
We will exclude ambiguous or uncategorized variants since we don't
have any in this data.

```{r calculate-mf-2}
mf_data_6_dose <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "dose_group",
  subtype_resolution = "base_6",
  variant_types = c("-ambiguous", "-uncategorized")
)
```

We will set the desired order of our dose groups using
factor levels. We will specify to our function that the
dose_group variable should be used to order the levels
in the plot.

```{r order-level-p}
# Set the desired order for the dose group:
mf_data_6_dose$dose_group <- factor(
  mf_data_6_dose$dose_group,
  levels = c("Control", "Low", "Medium", "High")
)
```

### Proportion

_Example 5. Plot the base_6 proportions for each dose group._

```{r plot-spectra-cap1,  include=FALSE}
caption <- paste(
  "Mutation spectrum (minimum) per Dose. Subtypes include single-nucleotide,",
  "variants at 6-base resolution, complex variants, deletions, insertions,",
  "multi-nucleotide variants (mnv) and structural variants (sv). Subtypes",
  "are represented by colour. Data is the proportion normalized to",
  "sequencing depth."
)
```

```{r plot-spectra-p, fig.cap=caption}
# Plot
plot <- plot_spectra(
  mf_data = mf_data_6_dose,
  group_col = "dose_group",
  subtype_resolution = "base_6",
  response = "proportion",
  group_order = "arranged",
  group_order_input = "dose_group",
  x_lab = "Dose Group",
  y_lab = "Subtype Proportion"
)
plot
```

### Frequency

_Example 6. Plot the base_6 frequences for each dose group._

```{r plot-spectra-cap2, include=FALSE}
caption <- paste(
  "Mutation spectrum (minimum) per Dose. Subtypes include single-nucleotide,",
  "variants at 6-base resolution, complex variants, deletions, insertions,",
  "multi-nucleotide variants (mnv) and structural variants (sv). Subtypes",
  "are represented by colour. Data is subtype frequency (mutations/bp)."
)
```

```{r plot-spectra-dose-f, fig.cap=caption}
# Plot
plot <- plot_spectra(
  mf_data = mf_data_6_dose,
  group_col = "dose_group",
  subtype_resolution = "base_6",
  response = "mf",
  group_order = "arranged",
  group_order_input = "dose_group",
  x_lab = "Dose Group",
  y_lab = "Subtype Frequency (mutations/bp)"
)
plot
```

### Sum

_Example 7. Plot the base_6 mutation sums for each dose group._

```{r plot-spectra-cap3,  include=FALSE}
caption <- paste(
  "Mutation spectrum (minimum) per Dose. Subtypes include single-nucleotide,",
  "variants at 6-base resolution, complex variants, deletions, insertions,",
  "multi-nucleotide variants (mnv) and structural variants (sv). Subtypes",
  "are represented by colour. Data is the mutation count."
)
```

```{r plot-spectra-dose-s}
# Plot
plot <- plot_spectra(
  mf_data = mf_data_6_dose,
  group_col = "dose_group",
  subtype_resolution = "base_6",
  response = "sum",
  group_order = "arranged",
  group_order_input = "dose_group",
  x_lab = "Dose Group",
  y_lab = "Subtype Mutation Count"
)
plot
```

#### Hierarchical Clustering

`plot_spectra()` integrates `cluster_spectra()` which performs unsupervised
hierarchical clustering of samples based on the mutation spectra.
`cluster_spectra()` uses `dist()` from the stats library to compute the
sample-to-sample distances using a user-defined distance measure (default
Euclidean). The resulting distance matrix is passed to `hclust()` to cluster
samples using the specified linkage method (default Ward). The function will
output a dendrogram visually representing the clusters' relationships and
hierarchy. The dendrogram will be overlaid on the `plot_spectra()` bar plot
and samples will be ordered accordingly.

_Example 8. Plot the base_6 mutation spectra per sample, with hierarchical_
_clustering. For this example we have created a new sample column with more_
intuitive sample names: new*sample_id. These names correspond to their\*
\_associated dose groups. We will see that samples largly cluster within their*
_dose groups._

```{r plot-spectra-cap4,  include=FALSE}
caption <- paste(
  "Mutation spectrum (minimum) per Sample. Subtypes include",
  "single-nucleotide variants at 6-base resolution, complex variants,",
  "deletions, insertions, multi-nucleotide variants (mnv) and structural",
  "variants (sv). Subtypes are represented by colour. Data is the proportion",
  "normalized to sequencing depth. Samples are clustered based on the",
  "Euclidean distance between their subtype proportions."
)
```

```{r plot-spectra-clustered, fig.wide=TRUE, fig.cap=caption}
# Calculate the mf data at the 6-base resolution for each sample
mf_data_6 <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "new_sample_id",
  subtype_resolution = "base_6",
  variant_types = c("-ambiguous", "-uncategorized")
)
# Plot
plot <- plot_spectra(
  mf_data = mf_data_6,
  group_col = "new_sample_id",
  subtype_resolution = "base_6",
  response = "proportion",
  group_order = "clustered",
  x_lab = "Sample",
  y_lab = "Subtype Proportion"
)
plot
```

### Trinucleotide Plots {.tabset}

The 96-base SNV mutation subtypes can be vizualised using `plot_trinucleotide()`.
This function creates a bar plot of the 96-base SNV spectrum for all levels of a
user-defined group. Data can represent subtype mutation count (`sum`), frequency
(`mf`), or `proportion`. Aesthetics are consistent with COSMIC trinucleotide
plots. Plots can be automatically saved to a specified output directory.

_Example 9. plot the base_96 mutation spectra proportions for each dose group._

We will first calculate the mf at the 96-based resolution per dose group.
For this example, we will focus only on SNV mutations.

```{r calc-mf-96}
# Calculate the mf data at the 96-base resolution for each dose
mf_data_96_dose <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "dose_group",
  subtype_resolution = "base_96",
  variant_types = "snv"
)
```

```{r trinuc-plot-caption, include=FALSE}
caption <- paste(
  "96-base trinucleotide spectra (minimum) per Dose.",
  "Bars are the proportion of SNV subtypes within their trinucleotide context",
  "normalized to the sequencing depth. ",
  "Bars are coloured based on SNV subtype. Data labels indicate the total",
  "number of mutations for each SNV subtype."
)
```

```{r plot-trinucleotide, fig.wide=TRUE, fig.show='hide'}
# Plot
plots <- plot_trinucleotide(
  mf_96 = mf_data_96_dose,
  group_col = "dose_group",
  response = "proportion",
  mf_type = "min",
  output_path = NULL
)
```

#### Control Dose

```{r trinuc-C, fig.wide=TRUE, echo=FALSE, fig.cap=caption}
plots[1]
```

#### Low Dose

```{r trinuc-L, fig.wide=TRUE, echo=FALSE, fig.cap=caption}
plots[2]
```

#### Medium Dose

```{r trinuc-M, fig.wide=TRUE, echo=FALSE, fig.cap=caption}
plots[3]
```

#### High Dose

```{r trinuc-H, fig.wide=TRUE, echo=FALSE, fig.cap=caption}
plots[4]
```

### Heatmaps

Another option for vizualizing the base-96 mutation spectra is
`plot_trinucleotide_heatmap()`. This function creates a heatmap of the
96-base SNV proportions. Plots can be facetted by additional grouping
variables. Heatmaps are useful for making comparisons between experimental
variables when information density becomes too high to represent using
traditional plots.

_Example 10. Plot the 96-base SNV spectrum for each sample, facetted_
_by dose group._

```{r heatmap-cap, include=FALSE}
caption <- paste(
  "96-base trinucleotide spectra (minimum) per Sample, facetted by Dose.",
  "Colour represents the proportion of SNV subtypes within their",
  "trinucleotide context normalized to the sequencing depth."
)
```

```{r plot-heatmap, fig.wide=TRUE, fig.cap=caption}
# Calculate the mf data at the 96-base resolution for each sample
mf_data_96 <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = "sample",
  subtype_resolution = "base_96",
  variant_types = "snv",
  retain_metadata_cols = "dose_group"
)
mf_data_96$dose_group <- factor(
  mf_data_96$dose_group,
  levels = c("Control", "Low", "Medium", "High")
)
# Plot
plot <- plot_trinucleotide_heatmap(
  mf_data = mf_data_96,
  group_col = "sample",
  facet_col = "dose_group"
)
plot
```

# References

# Appendix

## Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```
