---
title: "MutSeqR: Analysing Genomic Targets"
author:
  - name: Annette E. Dodge
    affiliation: Environmental Health Science and Research Bureau, Health Canada, Ottawa, ON, Canada.
  - name: Matthew J. Meier
    affilitation: Environmental Health Science and Research Bureau, Health Canada, Ottawa, ON, Canada.
    email: matthew.meier@hc-sc.gc.ca
package: MutSeqR
output:
  BiocStyle::html_document:
    toc: true
    toc_float: yes
    code_folding: show
    code_download: yes
    css: custom.css
---

```{r setup, include=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(DT)
library(htmltools)
```

In this vignette, we will provide examples on how to perform an analysis of mutation frequencies across a panel of genomic targets. We are using TwinStrand's [Mouse Mutagenesis Panel](https://github.com/twinstrandbio/twinstrandbio-reference-data/) as an example which consists of 20 2.4kb targets; one on each autosome, with the exeption of chromosome 1 which has 2 targets (chr1 & chr1.2).

The example data is taken from @leblanc-2022. This data consists of 24 mouse
bone marrow samples sequenced with Duplex Sequencing. Mice were exposed to
three doses of benzo[a]pyrene (BaP) alongside vehicle controls, n = 6. The goal
of this project was to quntify the mutagenic effects of BaP in mouse bone
marrow. Example data is retrieved from MutSeqRData, an ExperimentHub data
package available via Bioconductor. Example data is retrieved from the
ExperimentHub index (eh) through specific accessors.

# Load MutSeqR and Example Data

```{r load-lib}
library(MutSeqR)

library(ExperimentHub)
# load the index
eh <- ExperimentHub()
```

# Generalized Linear Modeling by Target

## Model by Target {.tabset}

We will measure the effect of genomic target and BaP dose on mutation frequency (MF) using a generalized linear mixed model using `model_mf()`. We will determine whether the MF of each BaP dose group is significantly different from the Control individually for all 20 genomic targets. In this model, dose group and target label will be our fixed effects. We include the interaction between the two fixed effects. Sample will be used as a random effect to control for repeated measurements.

First, load the example data. This mutation data has already undergone necessary filtering with `filter_mut` (demonstrated in the vignette; MutSeqR Introduction). We will then calculate MF for each sample and genomic locus, while retaining the dose column. The "label" column refers to the label of the genomic target.

```{r mf-rg}
example_data <- eh[["EH9861"]]

mf_data_rg <- calculate_mf(
  mutation_data = example_data,
  cols_to_group = c("sample", "label"),
  subtype_resolution = "none",
  retain_metadata_cols = "dose_group"
)
```

Next, create a contrasts table that compares each dose group back to the
control for each genomic locus:

```{r make-contrasts2}
combinations <- expand.grid(dose_group = unique(mf_data_rg$dose_group),
                            label = unique(mf_data_rg$label))
combinations <- combinations[combinations$dose_group != "Control", ]
combinations$col1 <- with(combinations, paste(dose_group, label, sep = ":"))
combinations$col2 <- with(combinations, paste("Control", label, sep = ":"))
contrasts2 <- combinations[, c("col1", "col2")]
```

Finally, run the model. As this model is more complex, we will improve
covergence by supplying the control argument, which is passed directly to
`lme4::glmer()`.

```{r model-mf-rg, fig.show='hide'}
model_by_target <- model_mf(mf_data = mf_data_rg,
  fixed_effects = c("dose_group", "label"),
  test_interaction = TRUE,
  random_effects = "sample",
  muts = "sum_min",
  total_count = "group_depth",
  contrasts = contrasts2,
  reference_level = c("Control", "chr1"),
  control = lme4::glmerControl(optimizer = "bobyqa",
                               optCtrl = list(maxfun = 2e5))
)
```

### Residuals Histogram

```{r hist-caption2, include=FALSE}
caption <- paste(
  "GLMM residuals of MFmin modelled as an effect of Dose and Genomic Target.",
  "x is pearson's residuals, y is frequency.",
  "Plotted to validate model assumptions. n = 24."
)
```

```{r model-rg-hist, echo=FALSE, fig.small=TRUE, fig.cap=caption}
model_data <- model_by_target$model_data
hist_data <- hist(model_data$residuals, plot = FALSE)
ylim_max <- max(hist_data$counts) + 1
plot <- hist(model_data$residuals,
  main = "Pearson Residuals",
  col = "yellow",
  ylim = c(0, ylim_max)
)
```

### Residuals QQ-plot

```{r qq-caption2, include=FALSE}
caption <- paste(
  "GLMM residuals of MFmin modelled as an effect of Dose and Genomic Target",
  "expressed as a quantile-quantile plot.",
  "Y is the pearson's residuals of the model in ascending order",
  "x is the quantiles of standard normal distribution for n of 24.",
  "Plotted to validate model assumptions."
)
```

```{r model-rg-qq, echo=FALSE, fig.small=TRUE, fig.cap=caption}
qqplot <- stats::qqnorm(model_data$residuals, main = "QQ Plot of Residuals")
stats::qqline(model_data$residuals, col = "red")
```

### Model Summary

```{r model-rg-summary}
model_by_target$summary
```

### ANOVA

```{r model-rg-anova}
model_by_target$anova
```

### Model Estimates

**Table 20.** model_by_target$point_estimates: Estimated Mean Mutation Frequency per Dose and Genomic Target.

```{r model-rg-est, echo=FALSE}
DT::datatable(
  model_by_target$point_estimates,
  options = list(
    columnDefs = list(
      list(targets = 1:4,
           render = DT::JS("function(data, type, row, meta) {
            return data.toExponential(2);
          }"))
    ),
    rowCallback = DT::JS("function(row, data, dataIndex) {
        $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }")
  )
)
```

### Pairwise Comparisons

**Table 21.** model_by_target$pairwise_comparisons.

```{r model-rg-comp, echo=FALSE}
DT::datatable(
  model_by_target$pairwise_comparisons,
  options = list(
    scrollX = TRUE,
    columnDefs = list(
      list(targets = c("p.value", "adj_p.value"),
        render = DT::JS("function(data, type, row, meta) { return data.toExponential(2); }")
      )
    ),
    rowCallback = DT::JS(
      "function(row, data, dataIndex) {
      $('td:eq(2)', row).css('text-align', 'right'); // Align the content of column 2 to the right
      }"
    )
  )
) %>%
  DT::formatRound(columns = c(1:3, 6:7), digits = 2)
```

### Plot Model by Dose and Target

```{r plot-model-cap2, include=FALSE}
caption <- paste(
  "Mean Mutation Frequency Minimum (mutations/bp) per Genomic Target and Dose",
  "estimated using a generalized linear mixed model. Error bars are the SEM.",
  "Symbols indicate significance differences (p < 0.05) between dose levels",
  "for individual genomic regions."
)
```

```{r plot-model-rg, fig.wide=TRUE, fig.cap=caption}
# Define the order of the genomic targets for the x-axis:
# We will order them from lowest to highest MF at the High dose.
label_order <- model_by_target$point_estimates %>%
  dplyr::filter(dose_group == "High") %>%
  dplyr::arrange(Estimate) %>%
  dplyr::pull(label)

# Define the order of the doses for the fill
dose_order <- c("Control", "Low", "Medium", "High")

plot <- plot_model_mf(
  model = model_by_target,
  plot_type = "bar",
  x_effect = "label",
  plot_error_bars = TRUE,
  plot_signif = TRUE,
  ref_effect = "dose_group",
  x_order = label_order,
  fill_order = dose_order,
  x_label = "Target",
  y_label = "Mutation Frequency (mutations/bp)",
  fill_label = "Dose",
  plot_title = "",
  custom_palette = c("#ef476f",
                     "#ffd166",
                     "#06d6a0",
                     "#118ab2")
)
# Rotate the x-axis labels for clarity using ggplot2 functions.
plot <- plot + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90))
plot
```

# Visualizae Mutations along Target Regions

`plot_lollipop()`

# Retrieve Sequences of Target regions

`get_seq()` will retrive raw nucleotide sequences for specified genomic
intervals. The function can retrieve sequences in one of two ways. Either
specify an installed BSgenome from which to retrieve sequences or retrieve
sequences through API request to the UCSC database based on supplied species
and genome assembly.

Supply regions with a filepath, data frame or GRanges object containing the
specified genomic intervals. TwinStrand's Mutagenesis Panels are stored in
package files and can easily be retrieved.

Supply the function with the appropriate BS_genome from which to retrieve
sequences. You can use the find_BS_genome() function to identify a BS genome
to use based on species and genome assembly. Be sure to install the BS genome
prior to running get_seq() with BiocManager().

To retrieve sequences from the UCSC data base, please input the species and
assembly genome to the species and genome parameters.

\*_BS_genome, species, or genome_
_do not need to be specified for the Mutagenesis Panels as this information_
_is stored_.

Sequences are returned within a _GRanges_ object.

_Example 2.1. Retrieve the sequences for our example's target panel,_
_TwinStrand's Mouse Mutagenesis Panel_

```{r get-seq}
regions_seq <- get_seq(regions = "TSpanel_mouse")
regions_seq
```

_Example 2.2. Retrieve sequences for a custom interval of regions. We will use_
_the Human Mutagenesis Panel as an example._

```{r get-seq-custom}
# We will load the TSpanel_human regions file to obtain
# an example list of regions.
human <- load_regions_file("TSpanel_human")
regions_seq <- get_seq(regions = human,
                       is_0_based_rg = FALSE,
                       BS_genome = find_BS_genome("human", "hg38"),
                       padding = 0)
regions_seq
```

_Example 2.3. Retrieve sequences using UCSC database._

```{r get-seq-ucsc}
regions_seq <- get_seq(regions = human,
                       is_0_based_rg = FALSE,
                       padding = 0,
                       species = "human",
                       genome = "hg38",
                       ucsc = TRUE)
```

Sequences can be exported as FASTA files with `write_reference_fasta()`.
Supply this function with the GRanges object containing the sequences of the
regions. Each one will be written to a single FASTA file. The FASTA file will
be saved to the `output_path`. If NULL, the file will be saved to the working
directory.

```{r write-ref-fasta, eval=FALSE}
write_reference_fasta(regions_seq, output_path = NULL)
```

# References

# Appendix

## Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```
